### 2019-12-05 manuscript/main.tex
started latex manuscript

\documentclass[letterpaper]{article}
\usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in]{geometry}
\usepackage{graphicx}
%\usepackage{times}      % Loads the Times-Roman Fonts
%\usepackage{mathptmx}   % Loads the Times-Roman Math Fonts
\usepackage{helvet}
\usepackage[sf,pagestyles]{titlesec}
\usepackage{lineno}
\usepackage{setspace}

\setlength{\parindent}{0em}
\setlength{\parskip}{1em}

\begin{document}

\doublespace

\section*{Robustness of phylogenetic analysis for detecting clusters of new HIV infections}

\textbf{Authors:}

August Guang$^1$, Mark Howison$^2$, Mia Coetzer$^3$, Lauren Ledingham$^3$, Matt D'Antuono$^3$,
Philip A. Chan$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^3$

\small $^1$ Computing and Information Services, Brown University, Providence, RI, USA

\small $^2$ Research Improving People's Lives, Providence, RI, USA

\small $^3$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA

\small $^4$ Division of Applied Mathematics, Brown University, Providence, RI, USA

\small $^5$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA

\linenumbers

\section*{Abstract}

\textbf{Background:} Phylogenetic analysis of HIV sequences obtained as part of clinical care is increasingly applied to detect clustering of new HIV infections and inform public health interventions to disrupt transmission. Conventional approaches summarize the within-host HIV diversity using only a single consensus sequence of only the HIV pol gene per individual, typically from Sanger or next-generation sequencing (NGS).

\textbf{Methods:} In all newly HIV diagnosed individuals in the first half of 2013 from the largest HIV center in Rhode Island, USA, we evaluate the robustness of the consensus approach and the potential benefit of considering full within-host HIV diversity, available via NGS, and of near whole HIV genome for phylogenetic inference. We compare Sanger and NGS-derived pol consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data for pol and near-whole HIV genomes.

\textbf{Results:} The space of phylogenies inferred from sampling is multi-modal, suggesting that a consensus-inferred phylogeny is not an appropriate summary of within-host variation. Cluster inference differs in phylogenies from consensus sequences in three clinically-relevant regions (prrt, int, env) versus the whole genome, and using Sanger versus NGS data.

\textbf{Discussion:} The choice of gene region and sequencing and summary methods affects the detection of clusters, and should be considered carefully in public health applications of phylogenetic analysis to disrupt HIV transmission.

\section*{Background}

\section*{Methods}

\section*{Results}

\section*{Discussion}

\bibliography{references}

\end{document}

### 2019-12-10 manuscript/main.tex
started adding references; added intro section and figures

@@
 \usepackage[letterpaper,top=1in,bottom=1in,left=1in,right=1in]{geometry}
+\usepackage[utf8]{inputenc}
 \usepackage{graphicx}
-%\usepackage{times}      % Loads the Times-Roman Fonts
-%\usepackage{mathptmx}   % Loads the Times-Roman Math Fonts
+\usepackage[T1]{fontenc}
+\usepackage{times}      % Loads the Times-Roman Fonts
+\usepackage{mathptmx}   % Loads the Times-Roman Math Fonts
 \usepackage{helvet}
-\usepackage[sf,pagestyles]{titlesec}
+\usepackage[sf,bf]{titlesec}
 \usepackage{lineno}
 \usepackage{setspace}
+\usepackage[style=nejm]{biblatex}
+\addbibresource{references.bib}
+\usepackage{caption}
+\captionsetup[figure]{labelfont={sf,bf}}
 
@@
 
-\doublespace
-
 \section*{Robustness of phylogenetic analysis for detecting clusters of new HIV infections}
-
-\textbf{Authors:}
 
@@
 
-\small $^1$ Computing and Information Services, Brown University, Providence, RI, USA
+$^1$ Computing and Information Services, Brown University, Providence, RI, USA
 
-\small $^2$ Research Improving People's Lives, Providence, RI, USA
+$^2$ Research Improving People's Lives, Providence, RI, USA
 
-\small $^3$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
+$^3$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
 
-\small $^4$ Division of Applied Mathematics, Brown University, Providence, RI, USA
+$^4$ Division of Applied Mathematics, Brown University, Providence, RI, USA
 
-\small $^5$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
+$^5$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
 
+\doublespace
 \linenumbers
-
 \section*{Abstract}
@@
 \textbf{Background:} Phylogenetic analysis of HIV sequences obtained as part of clinical care is increasingly applied to detect clustering of new HIV infections and inform public health interventions to disrupt transmission.
-Conventional approaches summarize the within-host HIV diversity using only a single consensus sequence of only the HIV pol gene per individual, typically from Sanger or next-generation sequencing (NGS).
+Conventional approaches summarize the within-host HIV diversity with a single consensus sequence per individual of only the \emph{pol} gene, obtained from Sanger or next-generation sequencing (NGS).
 
-\textbf{Methods:} In all newly HIV diagnosed individuals in the first half of 2013 from the largest HIV center in Rhode Island, USA, we evaluate the robustness of the consensus approach and the potential benefit of considering full within-host HIV diversity, available via NGS, and of near whole HIV genome for phylogenetic inference.
-We compare Sanger and NGS-derived pol consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data for pol and near-whole HIV genomes.
+\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals in the first half of 2013 at the largest HIV center in Rhode Island, USA.
+We compare Sanger and NGS-derived \emph{pol} and near-whole genome consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data.
 
-\textbf{Results:} The space of phylogenies inferred from sampling is multi-modal, suggesting that a consensus-inferred phylogeny is not an appropriate summary of within-host variation.
-Cluster inference differs in phylogenies from consensus sequences in three clinically-relevant regions (prrt, int, env) versus the whole genome, and using Sanger versus NGS data.
+\textbf{Results:} The space of phylogenies inferred through sampling is multi-modal, suggesting that a consensus-inferred phylogeny is not an appropriate summary of within-host variation.
+Cluster inference differs in phylogenies from consensus sequences from Sanger and NGS data, and across gene regions.
 
-\textbf{Discussion:} The choice of gene region and sequencing and summary methods affects the detection of clusters, and should be considered carefully in public health applications of phylogenetic analysis to disrupt HIV transmission.
+\textbf{Discussion:} The choice of sequencing and summarization methods affects the detection of clusters, and should be considered carefully in public health applications of phylogenetic analysis to disrupt HIV transmission.
 
@@
 
+Clinicians are generally interested in inferring transmission links between HIV-infected individuals to improve HIV treatment and prevention.
+In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
+The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was recently identified as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.
+
+While historically the phylogenetic inference power of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic and transmission cluster analyses, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
+In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.
+
+The increasing availability of NGS technology has led to longer and deeper sequencing of HIV, and data sets that cover nearly the whole genome with thousands or more reads at each site.
+Recent evidence suggests improvements in both phylogenetic analysis and cluster inference from near-whole genome HIV sequences obtained with NGS.
+For example, Yebra \emph{et al.} \parencite{yebra} found that the accuracy of phylogenetic reconstruction and cluster inference on simulated sequences improved with longer genomic regions (with the best accuracy from a gag-pol-env concatenation).
+Novitsky \emph{et al.} \parencite{novitsky} similarly studied the effects on cluster inference of using longer genomic regions from real near-whole genome Sanger sequences, and found that the proportion of sequences in clusters increased with longer sequence regions.
+
+Most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
+Accordingly, researchers often summarize the within-host variation present in NGS data sets with a consensus sequence.
+In the larger context of phylogenetic methods (e.g.
+beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}.
+In the context of HIV, tihs assumption is that the consensus sequence adequately captures most of the relevant information about the distribution of sequences that might be constructed from the within-host variation given the NGS data set.
+Some studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
+
+In this study, we examine whether this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first half of 2013 from the largest HIV center in Rhode Island, USA.
+We present a new analytical method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of phylogenetic analysis and cluster inference when using the consensus approach.
+
 \section*{Methods}
+
+Our study and data collection were approved by Lifespan's Institutional Review Board.
+We collected viral sequences from patients newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
+Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
+We obtained whole genome viral sequences for the 37 included individuals using both Sanger and NGS.
+Patient identifiers were removed and all analysis was conducted with de-identified sequence data.
+
+We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}.
+We start by aligning each individual's NGS reads to all reference database of all sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, using a variant of the \emph{hivmmer} pipeline \parencite{howison} that has been extended to support near-whole genome HIV data.
+Next, we convert the resulting multiple sequence alignment to a profile hidden Markov model (pHMM).
+Because pHMMs are probabilistic representations of alignments, we can sample genomic sequences from each individual's pHMM.
+
+sequences further allows us to estimate the probability of a molecular transmission	228
+cluster given the intra-patient variation.  To do so, we first infer a multiple sequence	229
+alignments and phylogenetic tree for each sampled genomic sequence. Then, we	230
+compute the bootstrap support [37] for each sampled tree and use a 99% bootstrap	231
+cutoff to identify transmission clusters.  Phylogenetic support value thresholds are the	232
+most common approach used to identify transmission clusters [9], hence our decision to	233
+apply it to our workflow.  The number of times a particular transmission cluster is	234
+identified in the set of phylogenetic tree samples divided by the number of samples is	235
+taken as the probability of that transmission cluster given the  intra-patient variation.	236
+For  the profile-sampling approach we use the well-established pHMM  alignment tool	243
+HMMER 3.1b2 [38] to build, align and sample from the pHMMs.
+
+For both approaches we estimate multiple alignments across individuals using mafft version 7.305b [39] and maximum-likelihood phylogenies with bootstrap replicates using RAxML version 8.2.10.
 
@@
 
+\subsection*{Profile sampling captures inter-host diversity}
+
+Profile sampling allows us to visualize the degree of polymorphism within a patient (Figure 1) and sample genomic sequences in proportion to the within-host variation.
+It can also build a consensus sequence for easy comparison of new approaches to established methods.
+By visualizing the degree of polymorphism, we can assess whether genomic variation exists at significantly high enough frequencies such that neither consensus genomes nor any other point estimator are sufficient summaries.
+
+Figure 1 shows substantial variation exists in the HIV profiles of all patients.
+Across	all patients, we found that 5.63 percent of sites in the HIV genome are polymorphic, with 11.4 percent of those sites in the gag region, 21.4 percent of those sites in the pol region, and 29.5 percent of those sites in the env region of the genome.
+Here we define polymorphism to mean that a particular site has greater than 1% variation.
+This falls within the range of prior studies on within-host variation \parencite{li, zanini}.
+This suggests that consensus genomes, and in particular Sanger sequencing of the pol region, discards variation that is informative for inferring the phylogeny.
+
+\subsection*{Phylogenetic estimates are sensitive to inter-host diversity}
+
+Given that previous studies on individuals infected with HIV have shown high intra-patient variation, and given that we found similar results in our set of data (Figure 1), we investigated its impact on downstream topological variation in the phylogenetic tree.
+The results show that trees inferred from the profile-sampled sequences do not cluster around any of the trees inferred from a genome point estimate (Figure 2).
+
+When and MDS plot is generated for only the profile-sampled trees (Figure 3), they appeared to form two clusters.
+This is likely due to zoom on the profile-sampled trees, as can be seen from the axes scales for Figure 2 and Figure 3.
+Supertrees of the two clusters reveal that most of the clades in the two subtrees are the same, with only patients out of 37 having a different topological placement on the two trees (Figure 4).
+This suggests that distinct clusters of phylogenetic trees exist given the intra-patient variation, but that the differences between them are not large.
+However, the consensus genome appears to be an inadequate estimate summary for the intra-patient variation, as can be seen in Figure 5.
+
+\subsection*{Inferred clusters differ by sequencing method and genomic region}
+
 \section*{Discussion}
 
-\bibliography{references}
+The true HIV transmission network is often unknown, but phylogenetic trees of patient viral sequences can provide information on transmission clusters that identify links between patients [1] as well as relevant features of transmission dynamics.
+[22, 23].
+Current phylogenetic approaches to transmission inference utilize a single summary consensus sequence per patient, which ignores the extensive intra-patient viral diversity [6, 18].
+
+\printbibliography
+
+\begin{figure}[p!]
+	\caption{Intra-patient genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for a patient) is highest in the env region for most individuals, and lies within the range of previously reported values.}
+	\centering
+	\includegraphics[width=\linewidth]{Figure1}
+	\label{fig1}
+\end{figure}
+
+\begin{figure}[p!]
+	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach show that the space of phylogenies inferred for the prrt and int regions are multi-modal.
+The phylogenies from consensus sequences (black dots) are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
+	\centering
+	\includegraphics[width=\linewidth]{Figure2}
+	\label{fig2}
+\end{figure}
+
+\begin{figure}[p!]
+	\caption{The total branch length in each of the profile-sampled phylogenies also varies.
+The phylogenies from consensus sequences (black dots) can lie at extreme values within these distribution, both when considering the lengths across all branches and the lengths across only the branches at the tips.}
+	\centering
+	\includegraphics[width=4in]{Figure3}
+	\label{fig3}
+\end{figure}
+
+\begin{figure}[p!]
+	\caption{Clusters (vertical colored bars) inferred from the phylogenies of NGS consensus sequences differ across genomic regions.
+The largest number of clusters was inferred from the int region, and the smallest number from the env region.
+Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
+	\centering
+	\includegraphics[width=\linewidth]{Figure4}
+	\label{fig4}
+\end{figure}
+
+%\begin{figure}[p!]
+%	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions.
+The largest number of clusters was inferred from the int region, and the smallest number from the env region.
+Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
+%	\centering
+%	\includegraphics{Figure5}
+%	\label{fig5}
+%\end{figure}
+%
+%\newpage
+
+\begin{figure}[p!]
+	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate bootstrapped cluster support from the profile sampling method.}
+	\centering
+	\includegraphics[width=4in]{Figure5}
+	\label{fig5}
+\end{figure}
 

### 2019-12-17 manuscript/main.tex
more work on methods and discussion

@@
 In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
-The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was recently identified as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.
+The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.
 
-While historically the phylogenetic inference power of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic and transmission cluster analyses, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
+While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
 In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.
@@
 beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}.
-In the context of HIV, tihs assumption is that the consensus sequence adequately captures most of the relevant information about the distribution of sequences that might be constructed from the within-host variation given the NGS data set.
-Some studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
+In the context of HIV, tihs assumption is that the consensus sequence adequately captures the relevant information about the distribution of sequences that might be constructed given the within-host variation measured in deep sequencing through NGS.
+Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
 
-In this study, we examine whether this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first half of 2013 from the largest HIV center in Rhode Island, USA.
-We present a new analytical method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of phylogenetic analysis and cluster inference when using the consensus approach.
+In this study, we examine how well this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA.
+We present a new analysis method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.
 
@@
 
+\subsection*{Data collection and sequencing}
+
 Our study and data collection were approved by Lifespan's Institutional Review Board.
-We collected viral sequences from patients newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
-Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
-We obtained whole genome viral sequences for the 37 included individuals using both Sanger and NGS.
+We collected viral sequences from 37 individuals newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
+Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during the first six months of 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
 Patient identifiers were removed and all analysis was conducted with de-identified sequence data.
 
+We obtained near-whole genome viral sequences for the 37 participants using both Sanger and NGS sequencing methods.
+Blood specimens were obtained from participants with their consent and processed to isolate peripheral blood mononuclear cells (PMBC), buffy coats, and plasma.
+From the collected plasma, total nucleic acid was extracted for genotyping.
+An in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}.
+For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV.
+Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and sequenced by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads.
+Sanger consensus sequences were generated manually using Sequencher version 5.2.4 (Gene Codes, Ann Arbor, MI) to confirm degenerate nucleotides.
+NGS data were processed and demultiplexed using BaseSpace cloud based application?
+(Illumina, San Diego, CA).
+
+\subsection*{Profile sampling}
+
 We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}.
-We start by aligning each individual's NGS reads to all reference database of all sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, using a variant of the \emph{hivmmer} pipeline \parencite{howison} that has been extended to support near-whole genome HIV data.
-Next, we convert the resulting multiple sequence alignment to a profile hidden Markov model (pHMM).
-Because pHMMs are probabilistic representations of alignments, we can sample genomic sequences from each individual's pHMM.
+We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0).
+Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to all group M reference sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, and produces a codon frequency table across the near-whole HIV genome, which we refer to as the individual's HIV \emph{profile}.
 
-sequences further allows us to estimate the probability of a molecular transmission	228
-cluster given the intra-patient variation.  To do so, we first infer a multiple sequence	229
-alignments and phylogenetic tree for each sampled genomic sequence. Then, we	230
-compute the bootstrap support [37] for each sampled tree and use a 99% bootstrap	231
-cutoff to identify transmission clusters.  Phylogenetic support value thresholds are the	232
-most common approach used to identify transmission clusters [9], hence our decision to	233
-apply it to our workflow.  The number of times a particular transmission cluster is	234
-identified in the set of phylogenetic tree samples divided by the number of samples is	235
-taken as the probability of that transmission cluster given the  intra-patient variation.	236
-For  the profile-sampling approach we use the well-established pHMM  alignment tool	243
-HMMER 3.1b2 [38] to build, align and sample from the pHMMs.
+We construct fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile.
+The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level.
+We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual.
+We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}.
+In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences.
+We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance.
+Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.
 
-For both approaches we estimate multiple alignments across individuals using mafft version 7.305b [39] and maximum-likelihood phylogenies with bootstrap replicates using RAxML version 8.2.10.
+In addition to performing these analyses on the near-whole genome sequences (``wgs''), we also perform them on subsets of the sequences in three clinicially relevant regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
+The prrt region and \emph{int} are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy.
+The \emph{env} region is...
+Rami, why is this routinely sequenced??
 
@@
 
-\subsection*{Profile sampling captures inter-host diversity}
+\subsection*{Profile sampling captures within-host diversity}
+
+We examine how much
 
@@
 
-The true HIV transmission network is often unknown, but phylogenetic trees of patient viral sequences can provide information on transmission clusters that identify links between patients [1] as well as relevant features of transmission dynamics.
-[22, 23].
-Current phylogenetic approaches to transmission inference utilize a single summary consensus sequence per patient, which ignores the extensive intra-patient viral diversity [6, 18].
+A limitation of our study is the small number of participants.
+The participants do have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria.
+However, the overall size of the HIV epidemic in Rhode Island is much larger, estimated as 2,396 individuals in 2016 \parencite{ridoh}.
+In future work, we hope to apply the profile sampling method to larger NGS data sets, to assess if cluster inference from Sanger versus NGS data agrees or differs especially for clusters larger than three individuals, which was the largest cluster size in our current data set.
+
+Our construction of HIV profiles from NGS data is limited by the accuracy of the NGS assays themselves.
+The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation.
+Sequencing protocols such as Primer ID \cite{jabara} have been introduced to reduce and correct for these biases.
+Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.
+
+The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \cite{fauci}.
+Current phylogenetic approaches do not fully utilize the information on within-host diversity available from deeper and longer NGS sequencing technologies.
+As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,
 

### 2019-12-30 lib/Figure1.R
Figure 1: tried adding viral load

@@ -1,8 +1,16 @@
 library(tidyverse)
+library(gridExtra)
 
 args <- commandArgs(trailingOnly=TRUE)
-distfiles <- args[1:(length(args)-1)]
-outfile  <- args[length(args)]
+
+samplefile <- args[1]
+distfiles  <- args[2:(length(args)-1)]
+outfile    <- args[length(args)]
+
+samples <- read_csv(samplefile) %>%
+           filter(Included==1) %>%
+           mutate(Dataset=factor(MC)) %>%
+           select("Dataset", "VL")
 
 Gene <- c()
 Dataset <- c()
@@ -28,7 +36,8 @@ for (distfile in distfiles)
 
 data <- tibble(Gene=factor(Gene, levels=c("prrt", "int", "env", "wgs")),
                Dataset=factor(Dataset),
-               Diversity=Diversity)
+               Diversity=Diversity) %>%
+        left_join(samples, by="Dataset")
 
 # Print min/max ranges for each region
 print(data %>%
@@ -39,11 +48,23 @@ print(data %>%
 env <- filter(data, Gene=="env") %>% arrange(Diversity)
 data <- mutate(data, Dataset=factor(Dataset, levels=env$Dataset))
 
-g <- ggplot(data, aes(x=Dataset, y=Diversity, colour=Gene)) +
-     geom_point() +
-     labs(x="Patient", y="Intra-patient Genetic Diversity") +
-     scale_y_continuous(labels=scales::percent_format(accuracy=1)) +
-     theme(legend.position="right",
-           axis.text.x=element_text(angle = 90))
+p1 <- ggplot(data, aes(x=Dataset, y=VL)) +
+      geom_bar(stat="identity") +
+      labs(x="Patient", y="Viral Load") +
+      theme(axis.text.x=element_text(angle=90),
+            axis.text=element_text(size=6),
+            axis.title=element_text(size=9))
 
-ggsave(outfile, g, width=12, height=3.5, units="in")
+p2 <- ggplot(data, aes(x=Dataset, y=Diversity, colour=Gene)) +
+      geom_point() +
+      labs(x="Patient", y="Within-host Genetic Diversity") +
+      scale_y_continuous(labels=scales::percent_format(accuracy=1)) +
+      theme(legend.position="bottom",
+            axis.text.x=element_text(angle=90),
+            axis.text=element_text(size=6),
+            axis.title=element_text(size=9),
[…]

### 2019-12-30 lib/Figure2.R

@@ -38,8 +38,8 @@ g <- ggplot(data, aes(x=X, y=Y)) +
      theme(legend.position="bottom",
            legend.title=element_blank(),
            axis.text=element_text(size=6),
-	   axis.title=element_text(size=9),
-	   strip.text=element_text(size=11),
-	   legend.text=element_text(size=11))
+           axis.title=element_text(size=9),
+           strip.text=element_text(size=11),
+           legend.text=element_text(size=11))
 
 ggsave(outfile, g, width=10, height=3.5, units="in")

### 2019-12-30 lib/Figure4.R
split Figure 4 into 4/5

@@ -1,8 +1,6 @@
 library(ape)
 library(tidyverse)
 library(ggtree)
-library(gridExtra)
-library(grid)
 
 args <- commandArgs(trailingOnly=TRUE)
 infiles <- args[1:(length(args)-1)]
@@ -67,7 +65,6 @@ for (i in seq(1, length(infiles), 2))
          geom_tiplab(size=2) +
          geom_treescale(y=-1) +
          xlim_tree(1.1*xscale) +
-         ggtitle(gene) +
          theme(plot.title=element_text(face="bold"))
     for (i in 1:nrow(support)) {
         cluster <- support$cluster[i]
@@ -76,14 +73,11 @@ for (i in seq(1, length(infiles), 2))
         }
         g <- g + geom_cladelabel(support$MRCA[i], label=round(support$support[i]), offset=0.1*xscale, color=colors[[cluster]])
     }
-    plots[[paste0(name, ":", gene)]] <- g
+    plots[[gene]] <- g
 }
 
 N <- length(plots) / 2
 
-pdf(outfile, width=18, height=12)
-grid.newpage()
-grid.draw(cbind(tableGrob(c("NGS\nConsensus", "Sanger\nConsensus"), theme=ttheme_minimal(base_size=16)),
-                arrangeGrob(grobs=plots, ncol=N, nrow=2),
-                size="last"))
+pdf(outfile, width=18, height=6)
+multiplot(plotlist=plots, ncol=4, labels=genes)
 dev.off()

### 2019-12-30 lib/FigureS1.R

@@ -10,15 +10,20 @@ print("Percent of variance explained by MDS axes:")
 print(round(mds$eig*100/sum(mds$eig), 1))
 
 data <- tibble(Gene=factor(genes, levels=c("prrt", "int", "env", "wgs")),
-               Consensus=consensus,
+               Name=names,
                X=mds$points[,1],
                Y=mds$points[,2]) %>%
-        arrange(Consensus)
+        arrange(Name)
 
-g <- ggplot(data, aes(x=X, y=Y, colour=Gene)) +
-     geom_point(shape=3) +
-     geom_point(data=filter(data, Consensus==TRUE), shape=16, colour="black") +
+g <- ggplot(data, aes(x=X, y=Y)) +
+     geom_point(aes(color=Gene), data=filter(data,  is.na(Name)), shape=3, show.legend=FALSE) +
+     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), color="black") +
      labs(x="MDS Axis 1", y="MDS Axis 2") +
-     theme(legend.position="bottom")
+     theme(legend.position="bottom",
+           legend.title=element_blank(),
+           axis.text=element_text(size=6),
+           axis.title=element_text(size=9),
+           strip.text=element_text(size=11),
+           legend.text=element_text(size=11))
 
-ggsave(outfile, g, width=12, height=12, units="in")
+ggsave(outfile, g, width=5, height=5, units="in")

### 2019-12-31 manuscript/main.tex
finished results section

@@
 While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
-In a recent meta study, Novitsky \emph{et al.} (forthcoming) found that 102 out of 107 published studies of HIV cluster inference between 2016 and 2019 used the \emph{pol} region alone.
+In a meta study of published studies on HIV cluster infereance, 98 out of 105 (93\%) analyzed the \emph{pol} region \cite{hassan}.
 
@@
 
-\subsection*{Profile sampling captures within-host diversity}
+\subsection*{Profile sampling estimates within-host diversity}
 
-We examine how much
+We estimate the within-host diversity for individuals as the average percent difference across all pairwise comparisons of their 500 profile-sampled nucleotide sequences.
+These pairwise differences are calculated using the Hamming distance \parencite{allam} (which is also called the $p$-distance in the literature on HIV genetic diversity \parencite{maldarelli, hassan}).
+Figure 1 shows the estimated percent diversity in each region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region.
+The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the largest range in estimated diversity (0.2\% to 4.0\%).
+The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs).
+These ranges are comparable to prior studies on within-host diversity \parencite{li, zanini}.
+These estimates of within-host diversity from the HIV profiles suggest that the consensus approach is discarding potentially-relevant information that is present in the NGS data.
 
-Profile sampling allows us to visualize the degree of polymorphism within a patient (Figure 1) and sample genomic sequences in proportion to the within-host variation.
-It can also build a consensus sequence for easy comparison of new approaches to established methods.
-By visualizing the degree of polymorphism, we can assess whether genomic variation exists at significantly high enough frequencies such that neither consensus genomes nor any other point estimator are sufficient summaries.
+\subsection*{Phylogenetic estimates are sensitive to within-host diversity}
 
-Figure 1 shows substantial variation exists in the HIV profiles of all patients.
-Across	all patients, we found that 5.63 percent of sites in the HIV genome are polymorphic, with 11.4 percent of those sites in the gag region, 21.4 percent of those sites in the pol region, and 29.5 percent of those sites in the env region of the genome.
-Here we define polymorphism to mean that a particular site has greater than 1% variation.
-This falls within the range of prior studies on within-host variation \parencite{li, zanini}.
-This suggests that consensus genomes, and in particular Sanger sequencing of the pol region, discards variation that is informative for inferring the phylogeny.
+Next, we investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates.
+To examine the variation in topology, we calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences.
+Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions (Figure 2).
+The results show that the topolical space is multi-model, and the consensus phylogenies inferred from the consensus sequences do not fully summarize the clustering that occurs in topological space.
+The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in difference clusters.
+There are two clusters in \emph{int}, and the NGS consensus phylogeny lies at the center of one of the clusters, while the Sanger consensus phylogeny is an outlier.
+The consensus phlyogenies lie at opposite ends of the primary MDS axis in \emph{env}, and of the secondary MDS axis in the wgs region.
 
-\subsection*{Phylogenetic estimates are sensitive to inter-host diversity}
+To examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches.
+We visualize the distribution of these branch length sums in Figure 3.
+Overall, the estimates are larger in \emph{env} and the wgs region, and smaller when restricting to only the tip branches.
+In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in Sanger consensus phylogeny for all branches in \emph{env}, or the NGS consensus phylogeny for tip branches in \emph{env}).
+In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the prrt region).
 
-Given that previous studies on individuals infected with HIV have shown high intra-patient variation, and given that we found similar results in our set of data (Figure 1), we investigated its impact on downstream topological variation in the phylogenetic tree.
-The results show that trees inferred from the profile-sampled sequences do not cluster around any of the trees inferred from a genome point estimate (Figure 2).
-
-When and MDS plot is generated for only the profile-sampled trees (Figure 3), they appeared to form two clusters.
-This is likely due to zoom on the profile-sampled trees, as can be seen from the axes scales for Figure 2 and Figure 3.
-Supertrees of the two clusters reveal that most of the clades in the two subtrees are the same, with only patients out of 37 having a different topological placement on the two trees (Figure 4).
-This suggests that distinct clusters of phylogenetic trees exist given the intra-patient variation, but that the differences between them are not large.
-However, the consensus genome appears to be an inadequate estimate summary for the intra-patient variation, as can be seen in Figure 5.
+Taken together, these results demonstrate that within-host diversity is associated with variation in phylogenetic estimates, and that the consensus approach provides only a point estimate that does not always adeuqatley summarize the underlying variation.
+For example, the variation can be multi-modal and the point estimates from the consensus phylogenies can be outliers.
 
@@
 
+Finally, we examine the clusters that are detected in the NGS vs.
+Sanger consensus phylogenies, and across the four regions.
+The NGS consensus phylogeny detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs.
+Each cluster appears as a colored bar in the phylograms in Figure 4.
+We use the proportion of times a cluster appears across the 500 profile-sampled phylogenies as a measure of support for that cluster, and report it next to the colored bar.
+Some clusters are robustly detected across all regions, such as the cluster (MC25, MC26, MC52), which has support between 98\% and 100\%.
+Other clusters are detected at lower support, and are not detected consistently across regions.
+For example, cluster (MC14, MC59) only appears in the wgs phylogeny, and at 62\% support.
+
+The Sanger consensus phylogeny detects 8 clusters in prrt, 7 in \emph{int}, 6 in \emph{env}, and 7 in wgs (Figure 5).
+As with the NGS consensus phylogeny, the support and consistency of cluster detection varies across regions.
+Cluster detection is mostly consistent with the NGS consensus, except for the cluster (MC17, MC21), which only occurs in \emph{env} in the Sanger consensus, and has zero support from the profile samples.
+Three clusters are consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28).
+
 \section*{Discussion}
+
+Because previous studies of the HIV epidemic have shown high intra-patient variation, 
+and because we estimate similar within-host diversity using HIV profiles of our NGS data,
 
@@
 The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation.
-Sequencing protocols such as Primer ID \cite{jabara} have been introduced to reduce and correct for these biases.
+Sequencing protocols such as Primer ID \parencite{jabara} have been introduced to reduce and correct for these biases.
 Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.
 
-The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \cite{fauci}.
-Current phylogenetic approaches do not fully utilize the information on within-host diversity available from deeper and longer NGS sequencing technologies.
+The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \parencite{fauci}.
+Current phylogenetic approaches do not fully utilize the information on within-host diversity available in near-whole genome NGS data.
 As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,

### 2019-12-31 manuscript/main.tex
finished discussion section

@@
 
-Clinicians are generally interested in inferring transmission links between HIV-infected individuals to improve HIV treatment and prevention.
+Clinicians are generally interested in inferring transmission links between HIV-infected individuals to inform and improve HIV treatment and prevention.
 In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
@@
 Recent evidence suggests improvements in both phylogenetic analysis and cluster inference from near-whole genome HIV sequences obtained with NGS.
-For example, Yebra \emph{et al.} \parencite{yebra} found that the accuracy of phylogenetic reconstruction and cluster inference on simulated sequences improved with longer genomic regions (with the best accuracy from a gag-pol-env concatenation).
+For example, Yebra \emph{et al.} \parencite{yebra} found that the accuracy of phylogenetic reconstruction and cluster inference on simulated sequences improved with longer genomic regions (with the best accuracy from a \emph{gag}-\emph{pol}-\emph{env} concatenation).
 Novitsky \emph{et al.} \parencite{novitsky} similarly studied the effects on cluster inference of using longer genomic regions from real near-whole genome Sanger sequences, and found that the proportion of sequences in clusters increased with longer sequence regions.
@@
 
-Finally, we examine the clusters that are detected in the NGS vs.
-Sanger consensus phylogenies, and across the four regions.
+Finally, we examine the clusters that are detected in the NGS versus Sanger consensus phylogenies, and across the four regions.
 The NGS consensus phylogeny detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs.
@@
 Other clusters are detected at lower support, and are not detected consistently across regions.
-For example, cluster (MC14, MC59) only appears in the wgs phylogeny, and at 62\% support.
+For example, cluster (MC14, MC59) only appears in the \emph{int} and wgs phylogenies, and at 60\% to 62\% support.
 
@@
 
-Because previous studies of the HIV epidemic have shown high intra-patient variation, 
-and because we estimate similar within-host diversity using HIV profiles of our NGS data,
+Our study provides new evidence on the utility of NGS HIV sequencing for phylogenetic analysis and cluster inference.
+The deeper sequencing provided by NGS can measure within-host diversity that is associated with variation in phylogenetic estimates.
+This variation is in turn associated with differences in cluster detection, which could have epidemiological consequences as public health officials increasingly look to cluster inference as a tool for informing and improving HIV prevention and treatment.
+Current approaches to HIV cluster inference almost exclusively use Sanger \emph{pol} consensus sequences, due to their availability from routine genotyping.
+In our results, this approach uncovers 8 clusters; however, only 3 of the 8 are robustly detected across regions and in both NGS and Sanger consensus sequences.
+
+In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in \emph{env} and wgs compared to prrt and \emph{int}.
+Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across regions.
+Some studies have found concordant clustering across \emph{gag}-\emph{env} \parencite{han} and \emph{gag}-\emph{pol}-\emph{env} \parencite{english,kaye}, while others found fewer clusters in \emph{pol} than in \emph{env} \parencite{kapaata}, or fewer clusters in \emph{gag}-\emph{env} than in \emph{pol} \parencite{ndiaye}.
+The additional information available in NGS data, along with the cluster support measures provided by profile sampling, may help resolve these differences.
+In our results, for example, the cluster support measure establishes four clusters with robust support ($\geq$95\%) in only prrt and \emph{int}: (MC37, MC53); (MC17, MC20, MC21); (MC41, MC47, MC56); and (MC45, MC58).
 
 A limitation of our study is the small number of participants.
-The participants do have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria.
-However, the overall size of the HIV epidemic in Rhode Island is much larger, estimated as 2,396 individuals in 2016 \parencite{ridoh}.
-In future work, we hope to apply the profile sampling method to larger NGS data sets, to assess if cluster inference from Sanger versus NGS data agrees or differs especially for clusters larger than three individuals, which was the largest cluster size in our current data set.
+The participants have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria.
+The overall size of the HIV epidemic in Rhode Island is much larger, estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented in this study.
+In future work, however, we hope to apply the profile sampling method to larger NGS data sets, to assess if cluster inference from Sanger versus NGS data agrees or differs especially for clusters larger than three individuals, which was the largest cluster size in this study.
 
@@
 Current phylogenetic approaches do not fully utilize the information on within-host diversity available in near-whole genome NGS data.
-As NGS data sets are increasingly available and more fully sample from the current HIV epidemic,
+As NGS data sets are increasingly available and become more representative of the current HIV epidemic, the additional information they measure has the potential to improve the robustness of cluster inference.
 
@@
 \begin{figure}[p!]
-	\caption{Intra-patient genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for a patient) is highest in the env region for most individuals, and lies within the range of previously reported values.}
+	\caption{Intra-patient genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for a patient) is highest in \emph{env} for most individuals, and lies within the range of previously reported values.}
 	\centering
@@
 	\centering
-	\includegraphics[width=4in]{Figure3}
+	\includegraphics[width=3.25in]{Figure3}
 	\label{fig3}
@@
 
-%\begin{figure}[p!]
-%	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions.
+\begin{figure}[p!]
+	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions.
 The largest number of clusters was inferred from the int region, and the smallest number from the env region.
 Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
-%	\centering
-%	\includegraphics{Figure5}
-%	\label{fig5}
-%\end{figure}
-%
-%\newpage
+	\centering
+	\includegraphics[width=\linewidth]{Figure5}
+	\label{fig5}
+\end{figure}
+
+\newpage
 
@@
 	\centering
-	\includegraphics[width=4in]{Figure5}
-	\label{fig5}
+	\includegraphics[width=3.25in]{Figure6}
+	\label{fig6}
 \end{figure}

### 2020-01-28 manuscript/main.tex
edits from August

@@
 
-August Guang$^1$, Mark Howison$^2$, Mia Coetzer$^3$, Lauren Ledingham$^3$, Matt D'Antuono$^3$,
-Philip A. Chan$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^3$
+August Guang$^{1,2}$, Mark Howison$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^6$
 
-$^1$ Computing and Information Services, Brown University, Providence, RI, USA
+$^1$ Center for Computational Biology of Human Disease, Brown University, Providence, RI, USA
 
-$^2$ Research Improving People's Lives, Providence, RI, USA
+$^2$ Center for Computation and Visualization, Brown University, Providence, RI, USA
 
-$^3$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
+$^3$ Research Improving People's Lives, Providence, RI, USA
 
@@
 $^5$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
+
+$^6$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
 
@@
 While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
-In a meta study of published studies on HIV cluster infereance, 98 out of 105 (93\%) analyzed the \emph{pol} region \cite{hassan}.
+In a meta study of published studies on HIV cluster inference, 98 out of 105 (93\%) analyzed the \emph{pol} region \parencite{hassan}.
 
@@
 
-Most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
-Accordingly, researchers often summarize the within-host variation present in NGS data sets with a consensus sequence.
+While the advantages of longer sequences in inferring transmission have been demonstrated, the advantages of deeper sequencing have not yet been fully investigated.
+In part this is because researchers often rely on consensus sequences, since most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
+Accordingly, researchers studying HIV transmission summarize the within-host HIV variation present in NGS data sets with a consensus sequence.
 In the larger context of phylogenetic methods (e.g.
 beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}.
-In the context of HIV, tihs assumption is that the consensus sequence adequately captures the relevant information about the distribution of sequences that might be constructed given the within-host variation measured in deep sequencing through NGS.
+In the context of HIV, this assumption is that a consensus sequence can adequately capture all of the relevant information about within-host variation available in a deeply-sequenced NGS data set.
 Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
 
-In this study, we examine how well this assumption underlying the consensus approach holds on a data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA.
-We present a new analysis method called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.
+In this study, we examine how well this assumption underlying the consensus approach holds on an NGS data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA.
+We present a new approach called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.
 
@@
 We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0).
-Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to all group M reference sequences from the Los Alamos National Lab HIV Database with the probabilistic multiple sequence aligner HMMER version 3.1b2 \parencite{eddy}, and produces a codon frequency table across the near-whole HIV genome, which we refer to as the individual's HIV \emph{profile}.
+A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences.
+Profile HMMs have been used in many kinds of biological sequence analysis and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}.
+Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database with the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome.
+We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.
 
@@
 We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual.
+We construct the 500 full-resolved sequences in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the proflie representation of the individual's within-host variation.
+
 We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}.
-In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences.
+In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences with the same tools and parameters.
 We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance.
@@
 
+\section*{Acknowledgments}
+
+Mia Coetzer, Lauren Ledingham, Matt D'Antuono, Philip A. Chan
+
 \printbibliography
@@
 \begin{figure}[p!]
-	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions. Numeric values indicate bootstrapped cluster support from the profile sampling method.}
+	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions.
+Numeric values indicate bootstrapped cluster support from the profile sampling method.
+A blank cell indicates that the cluster was not detected in that consensus sequence and genomic region.}
 	\centering

### 2020-02-09 manuscript/main.tex
disabled line numbers
edits from Casey

@@
 
-\section*{Robustness of phylogenetic analysis for detecting clusters of new HIV infections}
+\section*{Considering within-host diversity in phylogenetic analysis for detecting clusters of new HIV infections}
 
@@
 \doublespace
-\linenumbers
+%\linenumbers
 \section*{Abstract}
@@
 
-\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals in the first half of 2013 at the largest HIV center in Rhode Island, USA.
+\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island, USA, representing ~80\% of the state's HIV epidemic.
 We compare Sanger and NGS-derived \emph{pol} and near-whole genome consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data.
@@
 
-The increasing availability of NGS technology has led to longer and deeper sequencing of HIV, and data sets that cover nearly the whole genome with thousands or more reads at each site.
+The increasing availability of NGS technology has led to longer and deeper sequencing of HIV, and data sets routinely cover nearly the whole genome with thousands or more reads at each site.
 Recent evidence suggests improvements in both phylogenetic analysis and cluster inference from near-whole genome HIV sequences obtained with NGS.
@@
 While the advantages of longer sequences in inferring transmission have been demonstrated, the advantages of deeper sequencing have not yet been fully investigated.
-In part this is because researchers often rely on consensus sequences, since most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
+In part this is because researchers often rely on consensus sequences that discard all but the majority variant at each site.
+This is because most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
 Accordingly, researchers studying HIV transmission summarize the within-host HIV variation present in NGS data sets with a consensus sequence.
@@
 
-We construct fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile.
+For each patient, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile.
 The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level.
@@
 We estimate the within-host diversity for individuals as the average percent difference across all pairwise comparisons of their 500 profile-sampled nucleotide sequences.
+Variation in within-host diversity could be due to a variety of biological factors, such as virus mutation rate, effective viral population size, and time since infection, as well as technical factors such as sequencing depth and sequencing error rate.
 These pairwise differences are calculated using the Hamming distance \parencite{allam} (which is also called the $p$-distance in the literature on HIV genetic diversity \parencite{maldarelli, hassan}).
@@
 The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in difference clusters.
-There are two clusters in \emph{int}, and the NGS consensus phylogeny lies at the center of one of the clusters, while the Sanger consensus phylogeny is an outlier.
-The consensus phlyogenies lie at opposite ends of the primary MDS axis in \emph{env}, and of the secondary MDS axis in the wgs region.
+There are two clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier.
+The phlyogenies based on consensus sequences lie at opposite ends of the primary MDS axis in \emph{env}, and of the secondary MDS axis in the wgs region.
 
@@
 Overall, the estimates are larger in \emph{env} and the wgs region, and smaller when restricting to only the tip branches.
-In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in Sanger consensus phylogeny for all branches in \emph{env}, or the NGS consensus phylogeny for tip branches in \emph{env}).
+In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in phylogeny of Sanger consensus sequences for all branches in \emph{env}, or the phylogeny of NGS consensus sequences for tip branches in \emph{env}).
 In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the prrt region).
 
-Taken together, these results demonstrate that within-host diversity is associated with variation in phylogenetic estimates, and that the consensus approach provides only a point estimate that does not always adeuqatley summarize the underlying variation.
-For example, the variation can be multi-modal and the point estimates from the consensus phylogenies can be outliers.
+Taken together, these results demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
+For example, the variation can result in multi-modal phylogenetic results and the point estimates from the phylogenies of consensus sequences can be outliers.
 
@@
 
-Finally, we examine the clusters that are detected in the NGS versus Sanger consensus phylogenies, and across the four regions.
-The NGS consensus phylogeny detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs.
+Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four regions.
+The phylogeny of NGS consensus sequences detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs.
 Each cluster appears as a colored bar in the phylograms in Figure 4.
@@
 
-The Sanger consensus phylogeny detects 8 clusters in prrt, 7 in \emph{int}, 6 in \emph{env}, and 7 in wgs (Figure 5).
-As with the NGS consensus phylogeny, the support and consistency of cluster detection varies across regions.
+The phylogeny of Sanger consensus sequences detects 8 clusters in prrt, 7 in \emph{int}, 6 in \emph{env}, and 7 in wgs (Figure 5).
+As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varies across regions.
 Cluster detection is mostly consistent with the NGS consensus, except for the cluster (MC17, MC21), which only occurs in \emph{env} in the Sanger consensus, and has zero support from the profile samples.
@@
 
-Our study provides new evidence on the utility of NGS HIV sequencing for phylogenetic analysis and cluster inference.
-The deeper sequencing provided by NGS can measure within-host diversity that is associated with variation in phylogenetic estimates.
-This variation is in turn associated with differences in cluster detection, which could have epidemiological consequences as public health officials increasingly look to cluster inference as a tool for informing and improving HIV prevention and treatment.
-Current approaches to HIV cluster inference almost exclusively use Sanger \emph{pol} consensus sequences, due to their availability from routine genotyping.
-In our results, this approach uncovers 8 clusters; however, only 3 of the 8 are robustly detected across regions and in both NGS and Sanger consensus sequences.
+The deeper sequencing provided by NGS captures within-host diversity far more effectively than previous generations of data acquisition methods.
+The first step in many NGS studies, though, is to collapse this diversity into a single consensus sequence.
+This simplifies downstream analyses, but discards potentially interesting biological results and can mislead downstream analyses if the consensus is not a good representative of the total population of sequences.
+This could have epidemiological consequences as, for example, public health officials increasingly look to cluster inference as a tool for informing and improving HIV prevention and treatment.
+
+As data acquisition increasingly shifts to NGS approaches, it is important to compare new results to the Sanger \emph{pol} consensus sequences that have been used since the first molecular characterizations of HIV sequence diversity.
+Our analyses of Sanger \emph{pol} consensus sequences identifies 8 clusters; however, only 3 of the 8 are robustly detected across regions and in both NGS and Sanger consensus sequences.
 
@@
 In our results, for example, the cluster support measure establishes four clusters with robust support ($\geq$95\%) in only prrt and \emph{int}: (MC37, MC53); (MC17, MC20, MC21); (MC41, MC47, MC56); and (MC45, MC58).
+
+Our approach considered here allows for the consideration of within-host variation using standard-off the shelf tools, just by propogating variation further in the analyses.
+Rather than collapse variation into consensus sequences prior to phylogenetic analysis, we create a profile that captures that variation, perform multiple phylogenetic analyses on sequences drawn from this profile, and then summarize the phylogenetic analyses.
+XXX  Future tools could incorporate the uncertainty directly into the phylogenetic inference process.
 

### 2020-02-13 manuscript/main.tex
edits from Rami

@@
 
-\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island, USA, representing ~80\% of the state's HIV epidemic.
+\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island, USA, representing \~80\% of the state's HIV epidemic.
 We compare Sanger and NGS-derived \emph{pol} and near-whole genome consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data.
@@
 
-Clinicians are generally interested in inferring transmission links between HIV-infected individuals to inform and improve HIV treatment and prevention.
-In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
+Public health officials and providers are generally interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention.
+In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
 The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.
 
-While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was initially contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from routine clinical genotyping by commercial Sanger sequencing.
-In a meta study of published studies on HIV cluster inference, 98 out of 105 (93\%) analyzed the \emph{pol} region \parencite{hassan}.
+While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was suggested and contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from guideline-recommended routine clinical drug resistance testing by commercial Sanger sequencing \parencite{dhhs}.
+In a recent review of HIV cluster inference, 98 out of 105 (93\%) analyzed the \emph{pol} region \parencite{hassan}.
 
-The increasing availability of NGS technology has led to longer and deeper sequencing of HIV, and data sets routinely cover nearly the whole genome with thousands or more reads at each site.
+The increasing availability of NGS technology has led to longer and deeper sequencing of HIV, and data sets more routinely cover nearly the whole genome with thousands or more reads at each site \parencite{voelkerding}.
 Recent evidence suggests improvements in both phylogenetic analysis and cluster inference from near-whole genome HIV sequences obtained with NGS.
@@
 
-While the advantages of longer sequences in inferring transmission have been demonstrated, the advantages of deeper sequencing have not yet been fully investigated.
-In part this is because researchers often rely on consensus sequences that discard all but the majority variant at each site.
-This is because most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
+While the potential advantages of longer sequences in inferring transmission have been demonstrated, the advantages of deeper sequencing have not yet been fully investigated, and whether it can improve HIV molecular clustering inference is unknown.
+In part this is because researchers often rely on consensus sequences that discard all but the majority variant at each site, since most phylogenetic methods require a single fully resolved sequence for each individual included in the phylogeny.
 Accordingly, researchers studying HIV transmission summarize the within-host HIV variation present in NGS data sets with a consensus sequence.
@@
 
-In this study, we examine how well this assumption underlying the consensus approach holds on an NGS data set comprising all newly HIV-diagnosed individual in the first six months of 2013 from the largest HIV center in Rhode Island, USA.
+In this study, we examine how well the assumption underlying the consensus approach holds on a NGS data set comprising all newly HIV-diagnosed individuals during a six month time period from the largest HIV center in Rhode Island, USA, representing \~80\% of the state's HIV epidemic.
 We present a new approach called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.
@@
 
-Our study and data collection were approved by Lifespan's Institutional Review Board.
-We collected viral sequences from 37 individuals newly diagnosed with HIV during 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
-Inclusion criteria were: (i) HIV-infected adults, 18 years of age or older; (ii) diagnosed with HIV during the first six months of 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
+HIV-1 \emph{pol} sequences (HXB2 positions ) available for clinical care were collected from the 37 individuals newly diagnosed with HIV during the first half of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
+Inclusion criteria were: (i) adults with HIV, 18 years of age or older; (ii) diagnosed with HIV during the first six months of 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
 Patient identifiers were removed and all analysis was conducted with de-identified sequence data.
 
-We obtained near-whole genome viral sequences for the 37 participants using both Sanger and NGS sequencing methods.
-Blood specimens were obtained from participants with their consent and processed to isolate peripheral blood mononuclear cells (PMBC), buffy coats, and plasma.
-From the collected plasma, total nucleic acid was extracted for genotyping.
-An in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}.
+Blood specimens were obtained from participants with their consent and processed to isolate peripheral blood mononuclear cells (PMBC) and plasma.
+Using both Sanger and NGS, near-whole genome viral sequences were obtained from one of those compartments; plasma for participants with detectable viral load and PBMC for participants with undetectable viral load or unsuccessful plasma genotyping.
+Total nucleic acids were extracted for genotyping and an in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}.
 For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV.
@@
 A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences.
-Profile HMMs have been used in many kinds of biological sequence analysis and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}.
-Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database with the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome.
+Profile HMMs have been used in many kinds of biological sequence analyses and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}.
+Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database \parencite{losalamos} with the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome.
 We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.
@@
 We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual.
-We construct the 500 full-resolved sequences in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the proflie representation of the individual's within-host variation.
+We construct the 500 fully-resolved sequences in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the proflie representation of the individual's within-host variation.
 
@@
 
-In addition to performing these analyses on the near-whole genome sequences (``wgs''), we also perform them on subsets of the sequences in three clinicially relevant regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
+In addition to performing these analyses on the near-whole genome sequences (``wgs''), we also perform them on subsets of the sequences in three clinically relevant regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
 The prrt region and \emph{int} are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy.
-The \emph{env} region is...
-Rami, why is this routinely sequenced??
+The \emph{env} region is sequenced to genotypically infer viral tropism and co-receptor usage.
 
@@
 We estimate the within-host diversity for individuals as the average percent difference across all pairwise comparisons of their 500 profile-sampled nucleotide sequences.
-Variation in within-host diversity could be due to a variety of biological factors, such as virus mutation rate, effective viral population size, and time since infection, as well as technical factors such as sequencing depth and sequencing error rate.
+Variation in within-host diversity could be due to a variety of biological factors, such as virus mutation rate, effective viral population size, and time since infection, as well as technical factors such as sequencing depth and sequencing error rates.
 These pairwise differences are calculated using the Hamming distance \parencite{allam} (which is also called the $p$-distance in the literature on HIV genetic diversity \parencite{maldarelli, hassan}).
-Figure 1 shows the estimated percent diversity in each region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region.
-The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the largest range in estimated diversity (0.2\% to 4.0\%).
+Figure 1 shows the estimated within-host percent diversity in each examined genomic region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region.
+The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the overall largest range in estimated diversity (0.2\% to 4.0\%).
 The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs).
@@
 Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions (Figure 2).
-The results show that the topolical space is multi-model, and the consensus phylogenies inferred from the consensus sequences do not fully summarize the clustering that occurs in topological space.
-The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in difference clusters.
+The results show that the topological space is multi-modal, and the consensus phylogenies inferred from the consensus sequences do not fully summarize the clustering that occurs in topological space.
+The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in different clusters.
 There are two clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier.
@@
 
-Taken together, these results demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
+Taken together, the heterogeneity between the phylogenetic results from the profile sampling and the point estimates from the phylogenies of consensus sequences demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
 For example, the variation can result in multi-modal phylogenetic results and the point estimates from the phylogenies of consensus sequences can be outliers.
@@
 
-Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four regions.
-The phylogeny of NGS consensus sequences detects 8 clusters in prrt, 9 in \emph{int}, 4 in \emph{env}, and 6 in wgs.
+Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions.
+The phylogeny of NGS consensus sequences detected eight clusters in prrt, nine in \emph{int}, four in \emph{env}, and six in wgs.
 Each cluster appears as a colored bar in the phylograms in Figure 4.
@@
 
-The phylogeny of Sanger consensus sequences detects 8 clusters in prrt, 7 in \emph{int}, 6 in \emph{env}, and 7 in wgs (Figure 5).
+The phylogeny of Sanger consensus sequences detects eight clusters in prrt, seven in \emph{int}, six in \emph{env}, and seven in wgs (Figure 5).
 As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varies across regions.
@@
 The deeper sequencing provided by NGS captures within-host diversity far more effectively than previous generations of data acquisition methods.
-The first step in many NGS studies, though, is to collapse this diversity into a single consensus sequence.
-This simplifies downstream analyses, but discards potentially interesting biological results and can mislead downstream analyses if the consensus is not a good representative of the total population of sequences.
-This could have epidemiological consequences as, for example, public health officials increasingly look to cluster inference as a tool for informing and improving HIV prevention and treatment.
+Though the first step in many NGS studies, to collapse this diversity into a single consensus sequence simplifies downstream analyses, results presented here demonstrate that it discards potentially interesting biological results and can mislead downstream analyses, as if the consensus may not be a good representative of the total population of sequences.
+These data could have epidemiological consequences as, for example, HIV molecular clustering can trigger public health activities to inform and improve HIV prevention and treatment interventions \parencite{peters}.
+As data acquisition increasingly shifts to NGS approaches, it is important to compare new results to the Sanger \emph{pol} consensus sequences that have been used since the first molecular characterizations of HIV sequence diversity.
 
-As data acquisition increasingly shifts to NGS approaches, it is important to compare new results to the Sanger \emph{pol} consensus sequences that have been used since the first molecular characterizations of HIV sequence diversity.
 Our analyses of Sanger \emph{pol} consensus sequences identifies 8 clusters; however, only 3 of the 8 are robustly detected across regions and in both NGS and Sanger consensus sequences.
-
 In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in \emph{env} and wgs compared to prrt and \emph{int}.
@@
 
-Our approach considered here allows for the consideration of within-host variation using standard-off the shelf tools, just by propogating variation further in the analyses.
+Our approach presented here allows for the consideration of within-host variation using standard-off the shelf tools, just by propagating variation further in the analyses.
 Rather than collapse variation into consensus sequences prior to phylogenetic analysis, we create a profile that captures that variation, perform multiple phylogenetic analyses on sequences drawn from this profile, and then summarize the phylogenetic analyses.
@@
 
-A limitation of our study is the small number of participants.
+A limitation of our study is the small number of participants and the short time-frame patients were enrolled in.
 The participants have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria.
@@
 
-Our construction of HIV profiles from NGS data is limited by the accuracy of the NGS assays themselves.
+Our construction of HIV profiles from NGS data is also limited by the accuracy of the NGS assays themselves.
 The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation.
@@
 
-The true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \parencite{fauci}.
+In conclusion, the true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \parencite{fauci}.
 Current phylogenetic approaches do not fully utilize the information on within-host diversity available in near-whole genome NGS data.
-As NGS data sets are increasingly available and become more representative of the current HIV epidemic, the additional information they measure has the potential to improve the robustness of cluster inference.
+As NGS data sets are increasingly available and become more representative of the current HIV epidemic, we suggest that the additional information they measure has the potential to improve the robustness of cluster inference.
 
@@
 Mia Coetzer, Lauren Ledingham, Matt D'Antuono, Philip A. Chan
+
+R01AI136058, K24AI134359, P30AI042853, Brown University DEANs Award.
 

### 2020-04-27 lib/Figure1.R
updates to figure text size

@@ -40,10 +40,12 @@ env <- filter(data, Gene=="env") %>% arrange(Diversity)
 data <- mutate(data, Dataset=factor(Dataset, levels=env$Dataset))
 
 g <- ggplot(data, aes(x=Dataset, y=Diversity, colour=Gene)) +
-     geom_point() +
-     labs(x="Patient", y="Intra-patient Genetic Diversity") +
+     geom_point(size=2, shape=21, fill=NA) +
+     labs(x="Patient", y="Within-Host Diversity") +
      scale_y_continuous(labels=scales::percent_format(accuracy=1)) +
+     scale_colour_manual(values=c("black", "gray50", "red3", "#56B4E9")) +
+     theme_classic() +
      theme(legend.position="right",
-           axis.text.x=element_text(angle = 90))
+           axis.text.x=element_text(angle=90, vjust=0.5))
 
-ggsave(outfile, g, width=12, height=3.5, units="in")
+ggsave(outfile, g, width=6.5, height=2.5, units="in")

### 2020-04-27 lib/Figure2.R

@@ -29,17 +29,16 @@ data <- tibble(Gene=factor(Gene, levels=c("prrt", "int", "env", "wgs")),
         arrange(Name)
 
 g <- ggplot(data, aes(x=X, y=Y)) +
-     geom_point(aes(color=Gene), data=filter(data,  is.na(Name)), shape=3, show.legend=FALSE) +
-     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), color="black") +
+     geom_point(data=filter(data,  is.na(Name)), shape=3, color="gray30", size=1, alpha=0.1, show.legend=FALSE) +
+     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), size=3, color="black") +
      facet_grid(. ~ Gene) +
      xlim(-0.06, 0.06) +
      ylim(-0.06, 0.06) +
      labs(x="MDS Axis 1", y="MDS Axis 2") +
+     theme_minimal() +
      theme(legend.position="bottom",
            legend.title=element_blank(),
-           axis.text=element_text(size=6),
-           axis.title=element_text(size=9),
-           strip.text=element_text(size=11),
-           legend.text=element_text(size=11))
+           axis.text.x=element_text(angle=90, size=7, vjust=0.5, hjust=0.95),
+	   axis.text.y=element_text(size=7))
 
-ggsave(outfile, g, width=10, height=3.5, units="in")
+ggsave(outfile, g, width=6.5, height=3, units="in")

### 2020-04-27 lib/Figure3.R

@@ -51,16 +51,14 @@ data <- tibble(Gene=factor(Gene, levels=c("prrt", "int", "env", "wgs")),
                Tips=factor(Tips, levels=c("FALSE", "TRUE"), labels=c("All Branches", "Tip Branches")))
 
 g <- ggplot(data, aes(x=Gene, y=BranchLengths)) +
-     geom_violin(aes(fill=Gene), colour="white", show.legend=FALSE) +
-     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), colour="black") +
+     geom_violin(color=NA, fill="gray", show.legend=FALSE) +
+     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), size=3, colour="black") +
      expand_limits(y=0) +
      labs(y="Sum of Branch Lengths in Tree") +
      facet_grid(. ~ Tips) +
+     theme_minimal() +
      theme(legend.position="bottom",
            legend.title=element_blank(),
-           axis.text=element_text(size=6),
-           axis.title=element_text(size=9),
-           strip.text=element_text(size=11),
-           legend.text=element_text(size=11))
+	   panel.grid.major.x=element_blank())
 
-ggsave(outfile, g, width=5, height=5, units="in")
+ggsave(outfile, g, width=3.5, height=6.5, units="in")

### 2020-04-27 lib/Figure6.R

@@ -76,9 +76,6 @@ g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Cluster, label=Support)) +
      facet_grid(. ~ Name) + 
      theme(legend.position="none",
            panel.background=element_blank(),
-           axis.text=element_text(size=6),
-           axis.title=element_text(size=9),
-           strip.text=element_text(size=11),
-           legend.text=element_text(size=11))
+           strip.text=element_text(size=8))
 
-ggsave(outfile, g, width=5, height=5, units="in")
+ggsave(outfile, g, width=3.5, height=3.5, units="in")

### 2020-04-27 lib/codon-align-outgroup.py
rerun trees with HIV Group O outgroup

@@ -1,0 +1,52 @@
+import sys
+from Bio import Seq, SeqIO
+from collections import defaultdict
+from subprocess import Popen, PIPE
+
+in_file, outgroup_file, out_aa_file, out_nt_file = sys.argv[1:]
+
+# Read nucleotide sequences
+if in_file == "-":
+    nt_sequences = list(SeqIO.parse(sys.stdin, "fasta"))
+else:
+    with open(in_file, "r") as f:
+        nt_sequences = list(SeqIO.parse(f, "fasta"))
+
+# Append outgroup
+with open(outgroup_file, "r") as f:
+    nt_sequences += list(SeqIO.parse(f, "fasta"))
+
+# Translate to AA
+aa_sequences = "\n".join(">{}\n{}".format(record.id, record.seq.translate()) for record in nt_sequences)
+
+# Align with mafft
+with open(out_aa_file, "w") as f:
+    with Popen(["mafft", "-"], stdin=PIPE, stdout=f) as mafft:
+        mafft.communicate(bytes(aa_sequences, "ascii"))
+
+# Convert aligned AA to codons
+nt_sequences = dict((record.id, record.seq) for record in nt_sequences)
+ids = []
+columns = defaultdict(list)
+for record in SeqIO.parse(out_aa_file, "fasta"):
+    ids.append(record.id)
+    aa_seq = str(record.seq)
+    nt_seq = str(nt_sequences[record.id])
+    assert len(nt_seq) == 3*sum(1 for aa in aa_seq if aa != "-")
+    i = 0
+    for j, aa in enumerate(aa_seq):
+        if aa == "-":
+            columns[j].append("---")
+        else:
+            codon = nt_seq[i:i+3]
+            assert Seq.translate(codon) == aa.upper()
+            columns[j].append(codon)
+            i += 3
+assert ids[-1] == "outgroup"
+
+# Remove gapped columns
+N = len(nt_sequences) - 1
+idx = [i for i in sorted(columns) if columns[i].count("---") < N]
+with open(out_nt_file, "w") as f:
+    for j, id in enumerate(ids):
+        print(">{}\n{}".format(id, "".join(columns[i][j] for i in idx)), file=f)

### 2020-04-27 lib/raxml.sh

@@ -12,4 +12,4 @@ NAME=$6
 # create deterministic seed from alignment
 SEED=$(cat $FA | md5sum | grep -Po "\d" | tail -10 | tr -d "\n")
 
-raxmlHPC-PTHREADS-AVX -T $CPUS -s $FA -p $SEED -x $SEED -f a -N $N -m $MODEL -w $PWD/$OUTDIR -n $NAME
+raxmlHPC-PTHREADS-AVX -T $CPUS -s $FA -p $SEED -x $SEED -f a -N $N -m $MODEL -w $PWD/$OUTDIR -n $NAME -o outgroup

### 2020-04-27 manuscript/main.tex

@@
 
-\section*{Considering within-host diversity in phylogenetic analysis for detecting clusters of new HIV infections}
+\section*{Incorporating within-host diversity in phylogenetic analyses for detecting clusters of new HIV infections}
 
-August Guang$^{1,2}$, Mark Howison$^3$, Charles Lawrence$^4$, Casey W. Dunn$^5$, Rami Kantor$^6$
+August Guang$^{1,2}$, Mark Howison$^3$, Lauren Ledingham$^4$, Matthew D'Antuono$^4$, Philip A. Chan$^4$, Charles Lawrence$^5$, Casey W. Dunn$^6$, Rami Kantor$^4$
 
@@
 
-$^4$ Division of Applied Mathematics, Brown University, Providence, RI, USA
+$^4$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
 
-$^5$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
+$^5$ Division of Applied Mathematics, Brown University, Providence, RI, USA
 
-$^6$ Division of Infectious Diseases, The Alpert Medical School, Brown University, Providence, RI, USA
+$^6$ Department of Ecology and Evolutionary Biology, Yale University, New Haven, CT, USA
 
@@
 
-\textbf{Background:} Phylogenetic analysis of HIV sequences obtained as part of clinical care is increasingly applied to detect clustering of new HIV infections and inform public health interventions to disrupt transmission.
-Conventional approaches summarize the within-host HIV diversity with a single consensus sequence per individual of only the \emph{pol} gene, obtained from Sanger or next-generation sequencing (NGS).
+\textbf{Background:} Phylogenetic analyses of HIV sequences obtained in clinical care are increasingly applied to detect clustering of new HIV infections and inform public health interventions to disrupt transmission.
+Conventional approaches summarize within-host HIV diversity with only a single consensus sequence per individual of only the \emph{pol} gene, obtained from Sanger or next-generation sequencing (NGS).
 
-\textbf{Methods:} We evaluate the robustness of the consensus approach and the potential benefits of considering the within-host diversity in phylogenetic analysis and cluster inference for all newly HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island, USA, representing \~80\% of the state's HIV epidemic.
-We compare Sanger and NGS-derived \emph{pol} and near-whole genome consensus sequences to an alternate approach that samples many sequences per individual from a profile hidden Markov model of their NGS data.
+\textbf{Methods:} We evaluated the potential benefits of incorporating diversity into phylogenetic cluster inference for all newly-HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island.
+We compared Sanger- and NGS-derived \emph{pol} and near-whole genome consensus sequences to an alternate \emph{sample profiling} approach that incorporates deep-sequenced NGS data to capture within-host diversity.
 
-\textbf{Results:} The space of phylogenies inferred through sampling is multi-modal, suggesting that a consensus-inferred phylogeny is not an appropriate summary of within-host variation.
-Cluster inference differs in phylogenies from consensus sequences from Sanger and NGS data, and across gene regions.
+\textbf{Results:} The space of phylogenies inferred through \emph{profile sampling} reveals that consensus-inferred point estimates might not adequately represent them, both in topology and branch length.
+Cluster inference differed between Sanger- and NGS-derived sequences, and across gene regions; and only 30\% of clusters were detected unanimously.
 
-\textbf{Discussion:} The choice of sequencing and summarization methods affects the detection of clusters, and should be considered carefully in public health applications of phylogenetic analysis to disrupt HIV transmission.
+\textbf{Discussion:} \emph{Profile sampling} can incorporate within-host HIV diversity captured in NGS-derived deep sequencing into phylogenetic analyses.
+This additional information might improve the robustness of cluster detection and has the potential to aid in better understanding and disrupting transmission.
 
@@
 
-Public health officials and providers are generally interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention.
-In the absence of reliable patient contact histories, phylogenetic analysis of viral sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share a transmission link in the real, unobservable transmission network.
+Public health officials and providers are interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention \parencite{}.
+In the absence of reliable patient contact histories, phylogenetic analysis of HIV sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share an epidemiological link in the real, unobservable transmission network.
 The application of phylogenetic analysis and cluster inference techniques in public health interventions to disrupt transmission was delineated as one of the four key pillars for achieving the Department of Health and Human Services' recently-announced plan for ending the HIV epidemic in the US \parencite{fauci}.
 
-While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was suggested and contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from guideline-recommended routine clinical drug resistance testing by commercial Sanger sequencing \parencite{dhhs}.
+While historically the phylogenetic informativeness of the \emph{pol} region of the HIV genome was suggested and contested \parencite{hue, sturmer}, its use is now widespread in phylogenetic analysis and cluster inference, often due to the availability of \emph{pol} sequences from guideline-recommended routine clinical drug resistance testing, typically performed by commercial Sanger sequencing \parencite{dhhs}.
 In a recent review of HIV cluster inference, 98 out of 105 (93\%) analyzed the \emph{pol} region \parencite{hassan}.
@@
 For example, Yebra \emph{et al.} \parencite{yebra} found that the accuracy of phylogenetic reconstruction and cluster inference on simulated sequences improved with longer genomic regions (with the best accuracy from a \emph{gag}-\emph{pol}-\emph{env} concatenation).
-Novitsky \emph{et al.} \parencite{novitsky} similarly studied the effects on cluster inference of using longer genomic regions from real near-whole genome Sanger sequences, and found that the proportion of sequences in clusters increased with longer sequence regions.
+Novitsky \emph{et al.} \parencite{novitsky} similarly studied the effects on cluster inference of using longer genomic regions from real near-whole genome publicly available Sanger sequences, and found that the proportion of sequences in clusters increased with longer sequence regions.
 
@@
 Accordingly, researchers studying HIV transmission summarize the within-host HIV variation present in NGS data sets with a consensus sequence.
-In the larger context of phylogenetic methods (e.g.
-beyond their application to HIV sequences), this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}.
+Beyond the application of phylogenetic methods to HIV sequences, this consensus approach carries an underlying statistical assumption of \emph{low relative entropy} \parencite{guang}.
 In the context of HIV, this assumption is that a consensus sequence can adequately capture all of the relevant information about within-host variation available in a deeply-sequenced NGS data set.
-Some previous studies of HIV transmission dynamics have accounted for this variation with coalescent within-host evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
+Few previous studies of HIV transmission dynamics have accounted for within-host variation with coalescent evolutionary models \parencite{giardina, romero-severson}, but such models still assume a consensus sequence as the observed data.
 
-In this study, we examine how well the assumption underlying the consensus approach holds on a NGS data set comprising all newly HIV-diagnosed individuals during a six month time period from the largest HIV center in Rhode Island, USA, representing \~80\% of the state's HIV epidemic.
+In this study, we examine if and to what extent the incorporation of within-host variation available from deeply-sequenced NGS data improves consensus-sequence-based phylogenetic inference in a set comprising all newly HIV-diagnosed individuals during a six month time period from the Immunology Center, the largest HIV center in Rhode Island, USA.
 We present a new approach called \emph{profile sampling} that uses the within-host variation in the NGS data to assess the robustness of the consensus approach for phylogenetic analysis and cluster inference.
@@
 
-HIV-1 \emph{pol} sequences (HXB2 positions ) available for clinical care were collected from the 37 individuals newly diagnosed with HIV during the first half of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
-Inclusion criteria were: (i) adults with HIV, 18 years of age or older; (ii) diagnosed with HIV during the first six months of 2013; and (iii) available \emph{pol} sequence from routine drug resistance testing.
-Patient identifiers were removed and all analysis was conducted with de-identified sequence data.
+HIV-1 \emph{pol} Sanger sequences (HXB2 positions 2253-3554), available through clinical care, were collected from the 37 adults ($\geq$18 years) newly-diagnosed with HIV during the first six months of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
+Patients at this Center represent approximately 80% of the state's HIV epidemic.
 
-Blood specimens were obtained from participants with their consent and processed to isolate peripheral blood mononuclear cells (PMBC) and plasma.
-Using both Sanger and NGS, near-whole genome viral sequences were obtained from one of those compartments; plasma for participants with detectable viral load and PBMC for participants with undetectable viral load or unsuccessful plasma genotyping.
-Total nucleic acids were extracted for genotyping and an in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}.
+In addition, blood specimens were obtained from consenting participants and processed to isolate peripheral blood mononuclear cells (PMBC) and plasma.
+Using both Sanger sequencing and NGS, near-whole genome viral sequences were obtained from one of those compartments; plasma for participants with detectable viral load and PBMC for participants with undetectable viral load or unsuccessful plasma genotyping.
+Total nucleic acids were extracted fand an in-house genotyping assay was used to generate the near-whole genome based on previously published methods \parencite{nadai, di_giallonardo}.
 For each sample, two cDNA templates were generated by SuperscriptIII First Strand Synthesis System (Thermofisher, Carlsbad, CA), followed by eight separate nested PCR reactions; these eight amplicons span the near-whole genome of HIV.
-Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and sequenced by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads.
+Final amplicon products were sequenced by the Sanger method using 3100 Genetic Analyzer (Applied Biosystems, Foster City, CA) and by NGS using Nextera XT DNA Library Prep chemistry (Illumina, San Diego, CA) to generate multiplexed libraries for Illumina's MiSeq platform with 250 base paired-end reads.
 Sanger consensus sequences were generated manually using Sequencher version 5.2.4 (Gene Codes, Ann Arbor, MI) to confirm degenerate nucleotides.
-NGS data were processed and demultiplexed using BaseSpace cloud based application?
-(Illumina, San Diego, CA).
+NGS data were processed and demultiplexed using the BaseSpace cloud application (Illumina, San Diego, CA).
 
@@
 
-We introduce a new approach for incorporating within-host variation into phylogentic analysis, called \emph{profile sampling}.
+We introduce a new approach for incorporating within-host variation into phylogenetic analysis, called \emph{profile sampling}.
 We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0).
@@
 Profile HMMs have been used in many kinds of biological sequence analyses and are particularly well-suited to modeling variation in populations of sequences \parencite{eddy2004}.
-Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database \parencite{losalamos} with the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome.
+Briefly, hivmmer performs quality control and error correction in overlapping regions of the read pairs using PEAR version 0.9.11 \parencite{zhang}, translates them into each possible reading frame, aligns them in amino acid space to profile HMMs of all group M reference sequences from the Los Alamos National Lab HIV Database \parencite{losalamos} using the profile HMM alignment tool HMMER version 3.1b2 \parencite{eddy2013}, and produces a codon frequency table across the near-whole HIV genome.
 We refer to this resulting codon frequency table as the individual's HIV \emph{profile}.
 
-For each patient, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the profile.
+For each patient, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the \emph{profile}.
 The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level.
-We sample 500 sequences from each individual's profile, then collate the sampled sequences into 500 profile-sampled data sets, each having one sampled sequence per individual.
-We construct the 500 fully-resolved sequences in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the proflie representation of the individual's within-host variation.
+We then sample 500 sequences from each individual's \emph{profile}, and collate the sampled sequences into 500 \emph{profile-sampled} data sets, each having one sampled sequence per individual.
+We subsequently construct the 500 fully-resolved sequences per individual in order to use existing phylogenetic methods, because there is currently no published method to our knowledge for inferring a phylogeny directly from the profile representation of the individual's within-host variation.
 
-We perform phylogenetic inference on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}.
-In addition to the 500 profile-sampled phylogenies, we infer two additional phylogenies from the NGS consensus sequences and Sanger consensus sequences with the same tools and parameters.
+We estimate the within-host diversity for individuals as the average percent difference in nucleotides across all pairwise comparisons of their 500 profile-sampled nucleotide sequences.
+Variation in within-host diversity could be due to a variety of biological factors, such as virus mutation rate, effective viral population size, and time since infection, as well as technical factors such as sequencing depth and sequencing error rates.
+These pairwise differences are calculated using the Hamming distance \parencite{allam} (also called the $p$-distance in the HIV genetic diversity literature \parencite{maldarelli, hassan}).
+
+\subsection*{Phylogenetic inference}
+
+For profile sampling, we perform phylogenetic inference of near-whole genome sequences (wgs) on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}.
+In addition to profile sampling, we infer two phylogenies with the same tools and parameters, one from the NGS consensus sequences and one from Sanger consensus sequences.
 We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance.
@@
 
-In addition to performing these analyses on the near-whole genome sequences (``wgs''), we also perform them on subsets of the sequences in three clinically relevant regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
-The prrt region and \emph{int} are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy.
+Lastly, we also perform phylogenetic inference on three clinically relevant sub-genomic regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
+The prrt and \emph{int} regions are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy.
 The \emph{env} region is sequenced to genotypically infer viral tropism and co-receptor usage.
+
+We investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates.
+To examine the variation in topology, we first calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences.
+Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions.
+Next, to examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches, and visualize the distribution of these branch length sums.
+Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions.
+We use the proportion of times a cluster appears across the 500 \emph{profile-sampled} phylogenies as a measure of support for that cluster.
+We refer to this value as the \emph{cluster support} and note that it is similar to the use of bootstrap support for evaluating the robustness of an individual phylogeny's topology, but extends that idea to evaluating the robustness of cluster detection using within-host sequence variation.
 
@@
 
-\subsection*{Profile sampling estimates within-host diversity}
+\subsection*{Profile sampling estimates of within-host diversity}
 
-We estimate the within-host diversity for individuals as the average percent difference across all pairwise comparisons of their 500 profile-sampled nucleotide sequences.
-Variation in within-host diversity could be due to a variety of biological factors, such as virus mutation rate, effective viral population size, and time since infection, as well as technical factors such as sequencing depth and sequencing error rates.
-These pairwise differences are calculated using the Hamming distance \parencite{allam} (which is also called the $p$-distance in the literature on HIV genetic diversity \parencite{maldarelli, hassan}).
 Figure 1 shows the estimated within-host percent diversity in each examined genomic region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region.
@@
 The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs).
-These ranges are comparable to prior studies on within-host diversity \parencite{li, zanini}.
-These estimates of within-host diversity from the HIV profiles suggest that the consensus approach is discarding potentially-relevant information that is present in the NGS data.
+Such within-host estimations are not feasible via the conventional consensus Sanger or NGS approaches.
 
@@
 
-Next, we investigate the impact of within-host diversity on phylogenetic topology and evolutionary distance estimates.
-To examine the variation in topology, we calculate the pairwise geodesic distance \parencite{billera,owen} among the 500 phylogenies from the profile samples, as well as the phylogenies from the NGS and Sanger consensus sequences.
-Then we perform multi-dimensional scaling (MDS) on the resulting distance matrix to visualize the topological space in two dimensions (Figure 2).
-The results show that the topological space is multi-modal, and the consensus phylogenies inferred from the consensus sequences do not fully summarize the clustering that occurs in topological space.
-The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in different clusters.
-There are two clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier.
-The phlyogenies based on consensus sequences lie at opposite ends of the primary MDS axis in \emph{env}, and of the secondary MDS axis in the wgs region.
+Figure 2 demonstrates multi-dimensional scaling on the profile-sampled phylogenies and the phylogenies from the NGS and Sanger consensus sequences.
+The \emph{profile-sampling} approach reveals a multi-modal topological space in which the phylogenies inferred from the consensus sequences can be outliers.
+The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in different \emph{profile-sampled} clusters.
+There are two \emph{profile-sampled} clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier.
+The phylogenies based on consensus sequences lie at opposite ends of the primary \emph{profile-sampled} MDS axis in \emph{env}, and of the secondary \emph{profile-sampled} MDS axis in the wgs region.
 
-To examine the variation in estimated evolutionary distance, we sum the branch lengths within each phylogeny across all branches and across only the tip branches.
-We visualize the distribution of these branch length sums in Figure 3.
-Overall, the estimates are larger in \emph{env} and the wgs region, and smaller when restricting to only the tip branches.
+Figure 3 shows the distribution of branch length sums across the compared phylogenies.
+Overall, the estimates are larger in \emph{env} and the wgs regions, and smaller when restricting to only the tip branches.
 In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in phylogeny of Sanger consensus sequences for all branches in \emph{env}, or the phylogeny of NGS consensus sequences for tip branches in \emph{env}).
-In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the prrt region).
+In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the commonly-used prrt region).
 
-Taken together, the heterogeneity between the phylogenetic results from the profile sampling and the point estimates from the phylogenies of consensus sequences demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
+Taken together, the heterogeneity between the phylogenetic results from profile sampling and the consensus-inferred point estimates demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
 For example, the variation can result in multi-modal phylogenetic results and the point estimates from the phylogenies of consensus sequences can be outliers.
@@
 
-Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions.
-The phylogeny of NGS consensus sequences detected eight clusters in prrt, nine in \emph{int}, four in \emph{env}, and six in wgs.
-Each cluster appears as a colored bar in the phylograms in Figure 4.
-We use the proportion of times a cluster appears across the 500 profile-sampled phylogenies as a measure of support for that cluster, and report it next to the colored bar.
-Some clusters are robustly detected across all regions, such as the cluster (MC25, MC26, MC52), which has support between 98\% and 100\%.
-Other clusters are detected at lower support, and are not detected consistently across regions.
-For example, cluster (MC14, MC59) only appears in the \emph{int} and wgs phylogenies, and at 60\% to 62\% support.
+The phylogeny of NGS consensus sequences detected eight clusters in prrt, nine in \emph{int}, four in \emph{env}, and six in wgs (Figure 4).
+Some clusters were robustly detected across all regions (e.g.
+the cluster (MC25, MC26, MC52), which had support between 98\% and 100\%), while other clusters were detected at lower \emph{cluster support}, and were not detected consistently across regions (e.g.
+cluster (MC14, MC59), which only appeared in the \emph{int} and wgs phylogenies, with \emph{cluster support} of 60\% to 62\%).
 
-The phylogeny of Sanger consensus sequences detects eight clusters in prrt, seven in \emph{int}, six in \emph{env}, and seven in wgs (Figure 5).
-As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varies across regions.
-Cluster detection is mostly consistent with the NGS consensus, except for the cluster (MC17, MC21), which only occurs in \emph{env} in the Sanger consensus, and has zero support from the profile samples.
-Three clusters are consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28).
+The phylogeny of Sanger consensus sequences detected eight clusters in prrt, seven in \emph{int}, six in \emph{env}, and seven in wgs (Figure 5).
+As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varied across regions.
+Cluster detection was mostly consistent with the NGS consensus inference, except for the cluster (MC17, MC21), which only occurred in \emph{env} in the Sanger consensus, and had zero \emph{cluster support}.
+Three clusters were consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28).
+Only 30\% of clusters were detected unanimously.
+Figure 6 provides a visual summary of the differences in cluster detection between Sanger- and NGS-derived sequences and across genomic regions.
 
@@
 
-The deeper sequencing provided by NGS captures within-host diversity far more effectively than previous generations of data acquisition methods.
-Though the first step in many NGS studies, to collapse this diversity into a single consensus sequence simplifies downstream analyses, results presented here demonstrate that it discards potentially interesting biological results and can mislead downstream analyses, as if the consensus may not be a good representative of the total population of sequences.
-These data could have epidemiological consequences as, for example, HIV molecular clustering can trigger public health activities to inform and improve HIV prevention and treatment interventions \parencite{peters}.
-As data acquisition increasingly shifts to NGS approaches, it is important to compare new results to the Sanger \emph{pol} consensus sequences that have been used since the first molecular characterizations of HIV sequence diversity.
+In a dataset of all newly HIV-diagnosed individuals within six months at the largest HIV center in Rhode Island, USA, the deeper sequencing provided by NGS, represented by our introduced \emph{profile sampling}, captured within-host diversity more effectively than previous generations of data acquisition methods.
+These results impact phylogenetic inference across individuals and suggest that the consensus approach is discarding potentially-relevant information that is present in the NGS data.
+This additional information has the potential to improve the robustness of cluster detection.
 
-Our analyses of Sanger \emph{pol} consensus sequences identifies 8 clusters; however, only 3 of the 8 are robustly detected across regions and in both NGS and Sanger consensus sequences.
+The diversity ranges we estimated are comparable to prior studies on within-host diversity \parencite{li, zanini}.
+
+Although collapsing within-host diversity into a single consensus sequence simplifies downstream analyses, results presented here demonstrate that it discards potentially-relevant biological results and can mislead downstream resulting epidemiological consequences.
+For example, public health activities triggered by phylogenetic inference of HIV molecular clustering to inform and improve HIV prevention and treatment interventions can be affected \parencite{peters}.
+In our data, clusters vary in their \emph{profile-sampling} support, and consensus approaches can detect clusters that have low cluster support from deep-sequenced data.
+Public health activities may be better justified in scenarios where the clusters triggering them are determined to have high cluster support from deep-sequenced data.
+As data acquisition increasingly shifts to NGS approaches, it is important to compare results from new methods to the conventional Sanger \emph{pol} consensus sequences that have been used since the first molecular characterizations of HIV sequence diversity.
+
+Much of the enthusiasm about shifting from Sanger sequencing to NGS has been due to reducing costs and the ability to collect data on the entire HIV genome rather than a few genes.
+Our results suggest that much of the benefit of NGS may also be due to its greater depth of sequencing, which captures more variation in virus sequences within patients.
+This benefit can only be realized, though, if this variation is propagated to phylogenetic analyses, such as by the \emph{profile-sampling} method introduced here, rather than being collapsed to a consensus sequence, as is conventionally done.
+
+Our approach presented here allows for the consideration of within-host variation using standard off-the-shelf tools, just by propagating variation further in the analyses.
+Rather than collapse variation into consensus sequences prior to phylogenetic analysis, we suggest to create a profile that captures that variation, perform multiple phylogenetic analyses on sequences sampled from the profile, and then summarize the phylogenetic analyses.
+Future tools could incorporate the variation directly into the phylogenetic inference process itself \cite{guang}.
+
 In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in \emph{env} and wgs compared to prrt and \emph{int}.
-Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across regions.
+Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across genomic regions.
 Some studies have found concordant clustering across \emph{gag}-\emph{env} \parencite{han} and \emph{gag}-\emph{pol}-\emph{env} \parencite{english,kaye}, while others found fewer clusters in \emph{pol} than in \emph{env} \parencite{kapaata}, or fewer clusters in \emph{gag}-\emph{env} than in \emph{pol} \parencite{ndiaye}.
-The additional information available in NGS data, along with the cluster support measures provided by profile sampling, may help resolve these differences.
-In our results, for example, the cluster support measure establishes four clusters with robust support ($\geq$95\%) in only prrt and \emph{int}: (MC37, MC53); (MC17, MC20, MC21); (MC41, MC47, MC56); and (MC45, MC58).
+The additional information available in deep-sequenced NGS data, along with the cluster support measures provided by profile sampling, could help resolve these differences, as suggested here.
+In our data, not only were more clusters detected in the \emph{pol} region, but those clusters had higher cluster support as measured by deep-sequenced NGS data.
 
-Our approach presented here allows for the consideration of within-host variation using standard-off the shelf tools, just by propagating variation further in the analyses.
-Rather than collapse variation into consensus sequences prior to phylogenetic analysis, we create a profile that captures that variation, perform multiple phylogenetic analyses on sequences drawn from this profile, and then summarize the phylogenetic analyses.
-XXX  Future tools could incorporate the uncertainty directly into the phylogenetic inference process.
-
-A limitation of our study is the small number of participants and the short time-frame patients were enrolled in.
-The participants have a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, who met the inclusion criteria.
-The overall size of the HIV epidemic in Rhode Island is much larger, estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented in this study.
-In future work, however, we hope to apply the profile sampling method to larger NGS data sets, to assess if cluster inference from Sanger versus NGS data agrees or differs especially for clusters larger than three individuals, which was the largest cluster size in this study.
+One limitation of our study is the small number of participants and the short time-frame patients were enrolled in.
+The participants enrolled here represent a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, in which 80\% of the state's people with HIV are cared for.
+The overall size of the HIV epidemic in Rhode Island was estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented here.
+In future work, we will apply \emph{profile sampling} to larger NGS data sets, to assess cluster inference concordance between Sanger and NGS data, especially for clusters larger than three individuals, which was the largest cluster size here.
 
 Our construction of HIV profiles from NGS data is also limited by the accuracy of the NGS assays themselves.
-The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification during sample preparation.
-Sequencing protocols such as Primer ID \parencite{jabara} have been introduced to reduce and correct for these biases.
-Unfortunately, the Primer ID protocol supports only a limited region of the HIV genome, and has not yet been extended to support near-whole genome sequencing of HIV, but future work could study this limited region to determine if cluster inference changes with the implementation of a Primer ID protocol.
+The codon frequencies in the profiles may be biased measures of the true within-host diversity because of biases in PCR amplification.
+Sequencing protocols such as Primer ID \parencite{jabara} have been introduced to reduce and correct for these biases, and should be considered in the future.
 
-In conclusion, the true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials \parencite{fauci}.
-Current phylogenetic approaches do not fully utilize the information on within-host diversity available in near-whole genome NGS data.
-As NGS data sets are increasingly available and become more representative of the current HIV epidemic, we suggest that the additional information they measure has the potential to improve the robustness of cluster inference.
+In conclusion, the true HIV transmission network is unknown, but phylogenetic analysis and cluster inference are promising tools for aiding clinicians and public health officials in better understanding and in disrupting it \parencite{fauci}.
+Current phylogenetic approaches do not fully utilize the information on within-host diversity available in deep-sequenced, near-whole genome NGS data.
+As NGS data sets are increasingly available and become more representative of the HIV epidemics, we suggest that the additional information they measure has the potential to improve the robustness of HIV molecular cluster inference, the impact of which needs to be further investigated.
+
+\section*{Funding}
+
+This work was supported by the National Institutes of Health [R01AI136058, K24AI134359, P30AI042853, P20GM109035 ] and a Brown University DEANS Award.
 
@@
 
-Mia Coetzer, Lauren Ledingham, Matt D'Antuono, Philip A. Chan
-
-R01AI136058, K24AI134359, P30AI042853, Brown University DEANs Award.
+We thank Dr.
+Mia Coetzer for assistance with next-generation sequencing and feedback on an earlier draft of the manuscript.
+This research was conducted using computational resources and services at the Center for Computation and Visualization, Brown University.
 

### 2020-06-18 lib/Figure2.R
removed Figures S2-3 and added Figure 7

@@ -29,7 +29,7 @@ data <- tibble(Gene=factor(Gene, levels=c("prrt", "int", "env", "wgs")),
         arrange(Name)
 
 g <- ggplot(data, aes(x=X, y=Y, color=Gene)) +
-     geom_point(data=filter(data,  is.na(Name)), shape=3, color="gray30", size=1, alpha=0.1, show.legend=FALSE) +
+     geom_point(data=filter(data,  is.na(Name)), shape=3, size=3, alpha=0.1) +
      geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), size=3) +
      facet_grid(. ~ Gene) +
      xlim(-0.25, 0.25) +

### 2020-06-18 lib/Figure6.R

@@ -62,6 +62,6 @@ g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Consensus, label=Percent)) +
      scale_fill_manual(values=c("Both"="#ffffbf", "NGS Consensus"="#fc8d59", "Sanger Consensus"="#99d594", "Neither"="white")) +
      theme(legend.position="right",
            panel.background=element_blank(),
-           axis.text.x=element_text(angle=90))
+           axis.text.x=element_text(angle=90, vjust=0.5))
 
 ggsave(outfile, g, width=6.5, height=5, units="in")

### 2020-06-18 lib/FigureS1.R

@@ -15,9 +15,10 @@ data <- tibble(Gene=factor(genes, levels=c("prrt", "int", "env", "wgs")),
                Y=mds$points[,2]) %>%
         arrange(Name)
 
-g <- ggplot(data, aes(x=X, y=Y)) +
-     geom_point(aes(color=Gene), data=filter(data,  is.na(Name)), shape=3, show.legend=FALSE) +
-     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), color="black") +
+g <- ggplot(data, aes(x=X, y=Y, color=Gene)) +
+     geom_point(data=filter(data,  is.na(Name)), shape=3, alpha=0.2, size=3) +
+     geom_point(aes(shape=Name), data=filter(data, !is.na(Name)), size=3) +
+     scale_color_manual(values=c("#DF8F44", "#00A1D5", "#374E55", "#B24745")) +
      labs(x="MDS Axis 1", y="MDS Axis 2") +
      theme(legend.position="bottom",
            legend.title=element_blank(),
@@ -26,4 +27,4 @@ g <- ggplot(data, aes(x=X, y=Y)) +
            strip.text=element_text(size=11),
            legend.text=element_text(size=11))
 
-ggsave(outfile, g, width=5, height=5, units="in")
+ggsave(outfile, g, width=6.5, height=6.5, units="in")

### 2020-06-18 lib/FigureS2.R

@@ -1,37 +1,0 @@
-library(ape)
-library(tidyverse)
-library(ggtree)
-
-args <- commandArgs(trailingOnly=TRUE)
-treefiles <- args[1:(length(args)-1)]
-outfile <- args[length(args)]
-
-plots <- list()
-genes <- c()
-order <- c()
-
-for (treefile in treefiles)
-{
-    # Find gene name in filename path.
-    path <- str_split(treefile, "/")[[1]]
-    gene <- path[length(path)-1]
-    genes <- c(genes, gene)
-
-    # Load tree
-    trees <- read.tree(treefile)
-
-    if (length(order) == 0) {
-        tree <- ladderize(trees[[1]])
-        is_tip <- tree$edge[,2] <= length(tree$tip.label)
-        ordered_tips <- tree$edge[is_tip, 2]
-        order <- rev(tree$tip.label[ordered_tips])
-    }
-
-    # Plot tree
-    plots[[gene]] <- ggdensitree(trees, alpha=0.1*(100/length(trees)), tip.order=order) +
-                     geom_tiplab(size=1.6)
-}
-
-pdf(outfile, width=18, height=6)
-multiplot(plotlist=plots, ncol=4, labels=genes)
-dev.off()

### 2020-06-18 lib/Figure1.R
renumbered figures and changed output to emf format

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 distfiles <- args[1:(length(args)-1)]
@@ -52,4 +53,6 @@ g <- ggplot(data, aes(x=Dataset, y=Diversity, color=Gene, group=Gene)) +
            axis.text=element_text(color="black"),
            axis.text.x=element_text(angle=90, vjust=0.5))
 
-ggsave(outfile, g, width=6.5, height=5, units="in")
+emf(outfile, width=6.5, height=5)
+print(g)
+dev.off()

### 2020-06-18 lib/Figure2.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 mdsfiles <- args[1:(length(args)-1)]
@@ -45,4 +46,6 @@ g <- ggplot(data, aes(x=X, y=Y, color=Gene)) +
            axis.text.x=element_text(angle=90, size=7, vjust=0.5, hjust=0.95),
            axis.text.y=element_text(size=7))
 
-ggsave(outfile, g, width=6.5, height=3, units="in")
+emf(outfile, width=6.5, height=3)
+print(g)
+dev.off()

### 2020-06-18 lib/Figure3.R

@@ -1,5 +1,6 @@
 library(ape)
 library(tidyverse)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 filenames <- args[1:(length(args)-1)]
@@ -66,4 +67,6 @@ g <- ggplot(data, aes(x=Gene, y=BranchLengths, color=Gene)) +
            panel.grid.major.y=element_blank(),
            axis.text=element_text(color="black"))
 
-ggsave(outfile, g, width=6.5, height=4, units="in")
+emf(outfile, width=6.5, height=4)
+print(g)
+dev.off()

### 2020-06-18 lib/Figure4.R

@@ -1,83 +1,70 @@
 library(ape)
 library(tidyverse)
 library(ggtree)
+library(gridExtra)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 infiles <- args[1:(length(args)-1)]
 outfile <- args[length(args)]
 
-plots <- list()
-genes <- c()
-colors <- list()
-
-kelly_palette <- c(
-    red = "#be0032",
-    yellow = "#f3c300",
-    blue = "#0067a5",
-    olivegreen = "#2b3d26",
-    yellowgreen = "#8db600",
-    purplishpink = "#e68fac",
-    orange = "#f38400",
-    purple = "#875692",
-    reddishbrown = "#882d17",
-    green = "#008856",
-    buff = "#c2b280",
-    lightblue = "#a1caf1",
-    yellowishpink = "#f99379",
-    gray = "#848482",
-    yellowishbrown = "#654522",
-    reddishorange = "#e25822",
-    purplishred = "#b3446c",
-    greenishyellow = "#dcd300",
-    orangeyellow = "#f6a600",
-    violet = "#604e97"
-)
-
-for (i in seq(1, length(infiles), 2))
-{
-    treefile = infiles[i]
-    clusterfile = infiles[i+1]
-
-    # Find tree name in filename path.
-    if (endsWith(treefile, "consensus")) {
-        name <- "consensus"
-    } else if (endsWith(treefile, "sanger")) {
-        name <- "sanger"
-    }
+data <- bind_rows(lapply(infiles, function(infile) {
 
     # Find gene name in filename path.
-    path <- str_split(treefile, "/")[[1]]
+    path <- str_split(infile, "/")[[1]]
     gene <- path[length(path)-1]
-    genes <- c(genes, gene)
-
-    # Load tree
-    tree <- read.tree(treefile)
 
     # Load cluster support
-    support <- read_csv(clusterfile) %>% filter(get(name) == 1)
-    support$tips <- lapply(support$cluster, function(x) strsplit(x, ",")[[1]])
-    support$MRCA <- sapply(support$tips, function(x) getMRCA(tree, x))
-    print(support)
+    read_csv(infile) %>% mutate(Cluster=cluster,
+                                Gene=gene,
+                                Support=support,
+                                Consensus=case_when(consensus == 1 & sanger == 1 ~ "Both",
+                                                    consensus == 1               ~ "NGS Consensus",
[…]

### 2020-06-18 lib/Figure5.R

@@ -2,6 +2,7 @@ library(ape)
 library(tidyverse)
 library(ggtree)
 library(gridExtra)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 infiles <- args[1:(length(args)-1)]
@@ -56,12 +57,13 @@ data$Consensus <- factor(data$Consensus, levels=c("NGS Consensus", "Sanger Conse
 data$Percent <- paste0(formatC(data$Support, format="f", digits=1), "%")
 data$Support <- data$Support * 0.01
 
-g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Consensus, label=Percent)) +
+g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Support)) +
      geom_tile(color=NA) +
-     geom_text(size=3.3, color="black") +
-     scale_fill_manual(values=c("Both"="#ffffbf", "NGS Consensus"="#fc8d59", "Sanger Consensus"="#99d594", "Neither"="white")) +
+     scale_fill_gradient(name="Cluster Support %", low="#fee8c8", high="#e34a33", labels=scales::percent(0.25*0:4), breaks=0.25*0:4) +
      theme(legend.position="right",
            panel.background=element_blank(),
            axis.text.x=element_text(angle=90, vjust=0.5))
 
-ggsave(outfile, g, width=6.5, height=5, units="in")
+emf(outfile, width=6.5, height=5)
+print(g)
+dev.off()

### 2020-06-18 lib/FigureS1.R

@@ -1,4 +1,5 @@
 library(tidyverse)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 mdsfile <- args[1]
@@ -27,4 +28,6 @@ g <- ggplot(data, aes(x=X, y=Y, color=Gene)) +
            strip.text=element_text(size=11),
            legend.text=element_text(size=11))
 
-ggsave(outfile, g, width=6.5, height=6.5, units="in")
+emf(outfile, width=6.5, height=6.5)
+print(g)
+dev.off()

### 2020-06-18 lib/FigureS2.R

@@ -1,0 +1,83 @@
+library(ape)
+library(tidyverse)
+library(ggtree)
+
+args <- commandArgs(trailingOnly=TRUE)
+infiles <- args[1:(length(args)-1)]
+outfile <- args[length(args)]
+
+plots <- list()
+genes <- c()
+colors <- list()
+
+kelly_palette <- c(
+    red = "#be0032",
+    yellow = "#f3c300",
+    blue = "#0067a5",
+    olivegreen = "#2b3d26",
+    yellowgreen = "#8db600",
+    purplishpink = "#e68fac",
+    orange = "#f38400",
+    purple = "#875692",
+    reddishbrown = "#882d17",
+    green = "#008856",
+    buff = "#c2b280",
+    lightblue = "#a1caf1",
+    yellowishpink = "#f99379",
+    gray = "#848482",
+    yellowishbrown = "#654522",
+    reddishorange = "#e25822",
+    purplishred = "#b3446c",
+    greenishyellow = "#dcd300",
+    orangeyellow = "#f6a600",
+    violet = "#604e97"
+)
+
+for (i in seq(1, length(infiles), 2))
+{
+    treefile = infiles[i]
+    clusterfile = infiles[i+1]
+
+    # Find tree name in filename path.
+    if (endsWith(treefile, "consensus")) {
+        name <- "consensus"
+    } else if (endsWith(treefile, "sanger")) {
+        name <- "sanger"
+    }
+
+    # Find gene name in filename path.
+    path <- str_split(treefile, "/")[[1]]
+    gene <- path[length(path)-1]
+    genes <- c(genes, gene)
+
+    # Load tree
+    tree <- read.tree(treefile)
+
+    # Load cluster support
+    support <- read_csv(clusterfile) %>% filter(get(name) == 1)
+    support$tips <- lapply(support$cluster, function(x) strsplit(x, ",")[[1]])
+    support$MRCA <- sapply(support$tips, function(x) getMRCA(tree, x))
+    print(support)
+
+    # Plot tree
+    xscale <- max(node.depth.edgelength(tree))
+    g <- ggtree(tree) +
+         geom_tiplab(size=2) +
+         geom_treescale(y=-1) +
+         xlim_tree(1.1*xscale) +
+         theme(plot.title=element_text(face="bold"))
+    for (i in 1:nrow(support)) {
+        cluster <- support$cluster[i]
+        if (!(cluster %in% names(colors))) {
[…]

### 2020-06-18 lib/FigureS2.R
fixed Figures S2/S3

@@ -1,6 +1,7 @@
 library(ape)
 library(tidyverse)
 library(ggtree)
+library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
 infiles <- args[1:(length(args)-1)]
@@ -78,6 +79,6 @@ for (i in seq(1, length(infiles), 2))
 
 N <- length(plots) / 2
 
-pdf(outfile, width=18, height=6)
+emf(outfile, width=18, height=6)
 multiplot(plotlist=plots, ncol=4, labels=genes)
 dev.off()

### 2020-06-19 lib/Figure4.R
swapped Figures 4/5; increased size of Figures S2/S3

@@ -57,10 +57,9 @@ data$Consensus <- factor(data$Consensus, levels=c("NGS Consensus", "Sanger Conse
 data$Percent <- paste0(formatC(data$Support, format="f", digits=1), "%")
 data$Support <- data$Support * 0.01
 
-g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Consensus, label=Percent)) +
+g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Support)) +
      geom_tile(color=NA) +
-     geom_text(size=3.3, color="black") +
-     scale_fill_manual(values=c("Both"="#ffffbf", "NGS Consensus"="#fc8d59", "Sanger Consensus"="#99d594", "Neither"="white")) +
+     scale_fill_gradient(name="Cluster Support %", low="#fee8c8", high="#e34a33", labels=scales::percent(0.25*0:4), breaks=0.25*0:4) +
      theme(legend.position="right",
            panel.background=element_blank(),
            axis.text.x=element_text(angle=90, vjust=0.5))

### 2020-06-19 lib/Figure5.R

@@ -57,9 +57,10 @@ data$Consensus <- factor(data$Consensus, levels=c("NGS Consensus", "Sanger Conse
 data$Percent <- paste0(formatC(data$Support, format="f", digits=1), "%")
 data$Support <- data$Support * 0.01
 
-g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Support)) +
+g <- ggplot(data, aes(x=Gene, y=Cluster, fill=Consensus, label=Percent)) +
      geom_tile(color=NA) +
-     scale_fill_gradient(name="Cluster Support %", low="#fee8c8", high="#e34a33", labels=scales::percent(0.25*0:4), breaks=0.25*0:4) +
+     geom_text(size=3.3, color="black") +
+     scale_fill_manual(values=c("Both"="#ffffbf", "NGS Consensus"="#fc8d59", "Sanger Consensus"="#99d594", "Neither"="white")) +
      theme(legend.position="right",
            panel.background=element_blank(),
            axis.text.x=element_text(angle=90, vjust=0.5))

### 2020-06-19 lib/FigureS2.R

@@ -1,6 +1,8 @@
 library(ape)
 library(tidyverse)
 library(ggtree)
+library(gridExtra)
+library(grid)
 library(devEMF)
 
 args <- commandArgs(trailingOnly=TRUE)
@@ -66,19 +68,18 @@ for (i in seq(1, length(infiles), 2))
          geom_tiplab(size=2) +
          geom_treescale(y=-1) +
          xlim_tree(1.1*xscale) +
+         ggtitle(gene) +
          theme(plot.title=element_text(face="bold"))
     for (i in 1:nrow(support)) {
         cluster <- support$cluster[i]
         if (!(cluster %in% names(colors))) {
             colors[[cluster]] <- kelly_palette[length(colors)+1]
         }
-        g <- g + geom_cladelabel(support$MRCA[i], label=round(support$support[i]), offset=0.1*xscale, color=colors[[cluster]])
+        g <- g + geom_cladelabel(support$MRCA[i], label=round(support$support[i]), offset=0.1*xscale, color=colors[[cluster]], size=2)
     }
     plots[[gene]] <- g
 }
 
-N <- length(plots) / 2
-
-emf(outfile, width=18, height=6)
-multiplot(plotlist=plots, ncol=4, labels=genes)
+emf(outfile, width=6.5, height=8)
+grid.arrange(grobs=plots, ncol=2, nrow=2)
 dev.off()

### 2020-06-19 manuscript/main.tex
updated draft (2020-06-18)

@@
 \usepackage{caption}
+\usepackage{float}
 \captionsetup[figure]{labelfont={sf,bf}}
@@
 \textbf{Results:} The space of phylogenies inferred through \emph{profile sampling} reveals that consensus-inferred point estimates might not adequately represent them, both in topology and branch length.
-Cluster inference differed between Sanger- and NGS-derived sequences, and across gene regions; and only 30\% of clusters were detected unanimously.
+Cluster inference differed between Sanger- and NGS-derived sequences, and across gene regions.
+Only one of twelve of clusters was detected unanimously.
+Cluster support, however, was unanimously better when using the near-whole genome.
 
@@
 
-Public health officials and providers are interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention \parencite{}.
+Public health officials and providers are interested in inferring transmission links between individuals with HIV to inform and improve HIV treatment and prevention \parencite{hogben}.
 In the absence of reliable patient contact histories, phylogenetic analysis of HIV sequence data can and has been be used to infer transmission clusters \parencite{leitner}, under the assumption that two individuals sharing a most recent common ancestor in a phylogeny are more likely to share an epidemiological link in the real, unobservable transmission network.
@@
 HIV-1 \emph{pol} Sanger sequences (HXB2 positions 2253-3554), available through clinical care, were collected from the 37 adults ($\geq$18 years) newly-diagnosed with HIV during the first six months of 2013 and treated at The Miriam Hospital Immunology Center in Providence, Rhode Island, USA.
-Patients at this Center represent approximately 80% of the state's HIV epidemic.
+Patients at this Center represent approximately 80\% of the state's HIV epidemic.
 
@@
 NGS data were processed and demultiplexed using the BaseSpace cloud application (Illumina, San Diego, CA).
+Both Sanger and NGS consensus sequences were called at a 20\% threshold, meaning that any site with two or more nucleotides with $\geq 20\%$ frequency is represented by an ambiguous nucleotide code.
 
@@
 We introduce a new approach for incorporating within-host variation into phylogenetic analysis, called \emph{profile sampling}.
-We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (pending release as hivmmer version 0.3.0).
+We start by aligning each individual's NGS reads using the hivmmer pipeline \parencite{howison}, which we extended to support near-whole genome HIV data and to perform codon-aware alignment within each gene (hivmmer version 0.2.1).
 A key feature of this pipeline is its use of profile hidden Markov models (HMMs) to model and align collections of HIV sequences.
@@
 
-For each patient, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the \emph{profile}.
+For each individual, we construct a population of fully-resolved sequences by sampling codons at each site in the genome using the codon frequencies from the \emph{profile}.
 The collection of sampled sequences captures the empirical distribution of within-host variation at the codon level.
@@
 
-For profile sampling, we perform phylogenetic inference of near-whole genome sequences (wgs) on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with mafft version 7.313 \parencite{katoh} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}.
-In addition to profile sampling, we infer two phylogenies with the same tools and parameters, one from the NGS consensus sequences and one from Sanger consensus sequences.
-We perform cluster inference on these phylogenies using Cluster Picker \parencite{ragonnet-cronin} with thresholds of 80\% bootstrap support and 4.5\% genetic distance.
-Analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.
-
-Lastly, we also perform phylogenetic inference on three clinically relevant sub-genomic regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
+For profile sampling, we perform phylogenetic inference of near-whole genome sequences (wgs) on each of the 500 profile-sampled data sets by estimating a multiple sequence alignment with OMM\_MACSE version 10.02 \parencite{ranwez} and a maximum-likelihood phylogeny with the GTRCAT model and 100 rapid bootstrap replicates using RAxML version 8.2.12 \parencite{stamatakis}, with an HIV group O sequence (GenBank accession L20587.1) as the outgroup.
+We also perform this same phylogenetic inference on three clinically relevant sub-genomic regions: the protease and reverse transcriptase regions at the beginning of the \emph{pol} gene (``prrt''), the \emph{int} gene, and the \emph{env} gene.
 The prrt and \emph{int} regions are routinely sequenced in clinical care to detect drug resistance mutations and inform clinical choices of anti-retroviral therapy.
 The \emph{env} region is sequenced to genotypically infer viral tropism and co-receptor usage.
+
+In addition to profile sampling, we infer phylogenies with the same tools, regions and parameters for the NGS consensus sequences and the Sanger consensus sequences.
+We perform cluster inference on all phylogenies using Cluster Picker \parencite{ragonnet-cronin} with a threshold of 99\% bootstrap support.
+We do not impose a genetic distance threshold because empirically-justified thresholds that are comparable across the near-whole genome and the prrt, \emph{int}, and \emph{env} regions have not to our knowledge been established.
+This approach of using only a bootstrap criteria for cluster detection is consistent with methods commonly used in the broader literature on HIV cluster analysis \parencite{hassan}.
 
@@
 Finally, we examine the clusters that are detected in phylogenies of NGS consensus sequences versus Sanger consensus sequences, and across the four genomic regions.
-We use the proportion of times a cluster appears across the 500 \emph{profile-sampled} phylogenies as a measure of support for that cluster.
+We use the frequency that a cluster appears across the 500 \emph{profile-sampled} phylogenies as a measure of support for that cluster.
 We refer to this value as the \emph{cluster support} and note that it is similar to the use of bootstrap support for evaluating the robustness of an individual phylogeny's topology, but extends that idea to evaluating the robustness of cluster detection using within-host sequence variation.
+
+All analysis source code is available from \url{https://github.com/kantorlab/hiv-profile-sampling}.
 
@@
 Figure 1 shows the estimated within-host percent diversity in each examined genomic region across individuals, ordered by \emph{env}, which we expected \emph{a priori} to be the most variable region.
-The largest estimated diversity is in \emph{env} for individual MC28 (4.0\%), and \emph{env} has the overall largest range in estimated diversity (0.2\% to 4.0\%).
+The largest estimated diversity is in \emph{env} for individual MC28 (3.9\%), and \emph{env} has the overall largest range in estimated diversity (0.2\% to 3.9\%).
 The other regions have ranges of 0.2\% to 1.9\% (prrt), 0.1\% to 2.0\% (\emph{int}) and 0.2\% to 2.6\% (wgs).
@@
 
-Figure 2 demonstrates multi-dimensional scaling on the profile-sampled phylogenies and the phylogenies from the NGS and Sanger consensus sequences.
-The \emph{profile-sampling} approach reveals a multi-modal topological space in which the phylogenies inferred from the consensus sequences can be outliers.
-The prrt region displays four topological clusters, and the NGS and Sanger consensus phylogenies are located in different \emph{profile-sampled} clusters.
-There are two \emph{profile-sampled} clusters in \emph{int}, and the phylogeny of NGS consensus sequences lies at the center of one of the clusters, while the phylogeny of Sanger consensus sequences is an outlier.
-The phylogenies based on consensus sequences lie at opposite ends of the primary \emph{profile-sampled} MDS axis in \emph{env}, and of the secondary \emph{profile-sampled} MDS axis in the wgs region.
+Figure 2 demonstrates multi-dimensional scaling on the \emph{profile-sampled} phylogenies and the phylogenies from the NGS and Sanger consensus sequences within each genomic region (see Figure S1 for multi-dimensional scaling across all regions).
+The \emph{profile-sampling} approach reveals for each of the genomic regions a multi-modal topological space in which the phylogenies inferred from both the NGS and Sanger consensus sequences are outliers.
+A key different between the consensus and \emph{profile-sampled} sequences is that the consensus sequences contain ambiguous nucleotide codes at sites with two or more nucleotides with $\geq 20\%$ frequency.
+In contrast, \emph{profile-sampled} sequences by construction have no ambiguous sites, and ambiguity is instead incorporated into the analysis through the frequency of the ambiguous nucelotides across the 500 samples.
 
@@
 Overall, the estimates are larger in \emph{env} and the wgs regions, and smaller when restricting to only the tip branches.
-In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in phylogeny of Sanger consensus sequences for all branches in \emph{env}, or the phylogeny of NGS consensus sequences for tip branches in \emph{env}).
-In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in both consensus phylogenies for either all or tip branches in the commonly-used prrt region).
+In some cases, the consensus phylogenies provide an adequate summary of the distribution (as in the phylogeny of the NGS consensus sequence for tip branches for wgs).
+In other cases, the consensus phylogenies have estimates that are outliers in the distribution (as in the phylogenies from NGS and Sanger consensus sequences for all branches in wgs and \emph{env}).
 
-Taken together, the heterogeneity between the phylogenetic results from profile sampling and the consensus-inferred point estimates demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across patients, and that the consensus approach for collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
-For example, the variation can result in multi-modal phylogenetic results and the point estimates from the phylogenies of consensus sequences can be outliers.
+Taken together, the heterogeneity between the phylogenetic results from profile sampling and the consensus-inferred point estimates demonstrate that within-host virus sequence diversity impacts the inference of virus phylogeny across individuals, and that the consensus approach to handling ambiguity and collapsing within-host sequence variation can obscure both the magnitude and effect of these impacts.
 
@@
 
-The phylogeny of NGS consensus sequences detected eight clusters in prrt, nine in \emph{int}, four in \emph{env}, and six in wgs (Figure 4).
-Some clusters were robustly detected across all regions (e.g.
-the cluster (MC25, MC26, MC52), which had support between 98\% and 100\%), while other clusters were detected at lower \emph{cluster support}, and were not detected consistently across regions (e.g.
+Figure 4 illustrates the difference in cluster support across sequencing methods and genomic regions.
+Some clusters have consistently high support across all methods and regions, such as MC25/MC26/MC52 and MC14/MC59.
+Other clusters had higher support in certain genomic regions, such as MC17/MC20/MC21 in \emph{env} and wgs.
+One larger cluster, MC37/MC41/MC47/MC53/MC56, appeared only in the wgs region, and only about half of the time.
+Furthermore, the clusters detected in the wgs region had the highest overall clsuter support for every cluster.
+
+Figure 5 compares the differences in cluster detection between the phylogenies from the Sanger and the NGS consensus sequences.
+The phylogenies of NGS consensus sequences detected six clusters in prrt, seven in \emph{int}, eight in \emph{env}, and nine in wgs (Figure S2).
+Five clusters were robustly detected across all regions, while other clusters were detected at lower \emph{cluster support}, and were not detected consistently across regions (e.g.
 cluster (MC14, MC59), which only appeared in the \emph{int} and wgs phylogenies, with \emph{cluster support} of 60\% to 62\%).
 
-The phylogeny of Sanger consensus sequences detected eight clusters in prrt, seven in \emph{int}, six in \emph{env}, and seven in wgs (Figure 5).
-As with the phylogeny of NGS consensus sequences, the support and consistency of cluster detection varied across regions.
-Cluster detection was mostly consistent with the NGS consensus inference, except for the cluster (MC17, MC21), which only occurred in \emph{env} in the Sanger consensus, and had zero \emph{cluster support}.
-Three clusters were consistently detected across NGS and Sanger, and across all regions: (MC23, MC24); (MC25, MC26, MC52); and (MC27, MC28).
-Only 30\% of clusters were detected unanimously.
-Figure 6 provides a visual summary of the differences in cluster detection between Sanger- and NGS-derived sequences and across genomic regions.
+The phylogenies of Sanger consensus sequences detected fewer clusters than the phylogenies of NGS consensus sequence in all regions except \emph{env}.
+The phylogenies of Sanger consensus sequences detected four clusters in prrt, six in \emph{int}, nine in \emph{env}, and eight in wgs (Figure S3).
+Only one cluster (MC25/MC26/MC52) was consistently detected across the phylogenies from NGS and Sanger consensus sequences, and across all regions.
+Neither detected the larger cluster, MC37/MC41/MC47/MC53/MC56, which was detected only through profile sampling.
 
@@
 This additional information has the potential to improve the robustness of cluster detection.
-
-The diversity ranges we estimated are comparable to prior studies on within-host diversity \parencite{li, zanini}.
 
@@
 Much of the enthusiasm about shifting from Sanger sequencing to NGS has been due to reducing costs and the ability to collect data on the entire HIV genome rather than a few genes.
-Our results suggest that much of the benefit of NGS may also be due to its greater depth of sequencing, which captures more variation in virus sequences within patients.
+Our results suggest that much of the benefit of NGS may also be due to its greater depth of sequencing, which captures more variation in virus sequences within individuals.
 This benefit can only be realized, though, if this variation is propagated to phylogenetic analyses, such as by the \emph{profile-sampling} method introduced here, rather than being collapsed to a consensus sequence, as is conventionally done.
@@
 
-In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in \emph{env} and wgs compared to prrt and \emph{int}.
+In our comparison of cluster inference across genomic regions, we found that fewer clusters were detected overall in prrt and \emph{int} compared to \emph{env} and wgs.
 Prior studies of clustering from Sanger consensus sequences present mixed results on the prevalence of clustering across genomic regions.
@@
 The additional information available in deep-sequenced NGS data, along with the cluster support measures provided by profile sampling, could help resolve these differences, as suggested here.
-In our data, not only were more clusters detected in the \emph{pol} region, but those clusters had higher cluster support as measured by deep-sequenced NGS data.
+In our data, not only were more clusters detected in the near-whole length genome, but those clusters also had higher cluster support as measured by deep-sequenced NGS data.
 
-One limitation of our study is the small number of participants and the short time-frame patients were enrolled in.
+One limitation of our study is the small number of participants and the short timeframe they were enrolled in.
 The participants enrolled here represent a dense temporal sampling, and comprise all newly HIV-diagnosed individuals in a six month period at the largest HIV center in Rhode Island, USA, in which 80\% of the state's people with HIV are cared for.
 The overall size of the HIV epidemic in Rhode Island was estimated as 2,396 individuals in 2016 \parencite{ridoh}, but NGS data for this population are not currently available beyond those presented here.
-In future work, we will apply \emph{profile sampling} to larger NGS data sets, to assess cluster inference concordance between Sanger and NGS data, especially for clusters larger than three individuals, which was the largest cluster size here.
+In future work, we will apply \emph{profile sampling} to larger NGS data sets, to assess cluster inference concordance between Sanger and NGS data, especially for clusters larger than five individuals, which was the largest cluster size here.
 
@@
 
-This work was supported by the National Institutes of Health [R01AI136058, K24AI134359, P30AI042853, P20GM109035 ] and a Brown University DEANS Award.
+This work was supported by the National Institutes of Health [R01AI136058, K24AI134359, P30AI042853, P20GM109035] and a Brown University DEANS Award.
 
@@
 
-\begin{figure}[p!]
-	\caption{Intra-patient genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for a patient) is highest in \emph{env} for most individuals, and lies within the range of previously reported values.}
+\begin{figure}[H]
+	\caption{Intra-host genetic diversity (defined as the average percent difference across all pairwise comparisons of the 500 profile-sampled nucleotide sequences for an individual) is highest in \emph{env} for most individuals, and lies within the range of previously reported values.}
 	\centering
-	\includegraphics[width=\linewidth]{Figure1}
-	\label{fig1}
+	\includegraphics[width=6.5in]{Figure1.emf}
 \end{figure}
 
-\begin{figure}[p!]
-	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach show that the space of phylogenies inferred for the prrt and int regions are multi-modal.
-The phylogenies from consensus sequences (black dots) are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
+\begin{figure}[H]
+	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach show that the space of inferred phylogenies is multi-modal for all genomic regions.
+The phylogenies from consensus sequences (dot and triangle) are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
 	\centering
-	\includegraphics[width=\linewidth]{Figure2}
-	\label{fig2}
+	\includegraphics[width=6.5in]{Figure2.emf}
 \end{figure}
 
-\begin{figure}[p!]
+\begin{figure}[H]
 	\caption{The total branch length in each of the profile-sampled phylogenies also varies.
-The phylogenies from consensus sequences (black dots) can lie at extreme values within these distribution, both when considering the lengths across all branches and the lengths across only the branches at the tips.}
+The phylogenies from consensus sequences (dot and triangle) can lie at extreme values within these distribution, both when considering the lengths across all branches and the lengths across only the branches at the tips.}
 	\centering
-	\includegraphics[width=3.25in]{Figure3}
-	\label{fig3}
+	\includegraphics[width=6.5in]{Figure3.emf}
 \end{figure}
 
-\begin{figure}[p!]
+\begin{figure}[H]
+	\caption{The cluster support of clusters identified by profile sampling across genomic regions. A blank cell indicates that the cluster was not detected in that genomic region.}
+	\centering
+	\includegraphics[width=6.5in]{Figure4.emf}
+\end{figure}
+
+\begin{figure}[H]
+	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions.
+Numeric values indicate cluster support from the profile sampling method.
+A blank cell indicates that the cluster was not detected in that genomic region.}
+	\centering
+	\includegraphics[width=6.5in]{Figure5.emf}
+\end{figure}
+
+\begin{figure}[H]
+	\caption{Multi-dimensional scaling of pairwise geodesic distance among maximum-likelihood phylogenies from the profile-sampling approach shows that the space of inferred phylogenies are multi-modal for all genomic regions.
+The phylogenies from consensus sequences are point estimates that do not capture the full variation in phylogenies that can be inferred from deeply-sequenced NGS data.}
+	\centering
+	\includegraphics[width=6.5in]{FigureS1.emf}
+\end{figure}
+
+\begin{figure}[H]
 	\caption{Clusters (vertical colored bars) inferred from the phylogenies of NGS consensus sequences differ across genomic regions.
-The largest number of clusters was inferred from the int region, and the smallest number from the env region.
+The largest number of clusters was inferred from the wgs region, and the smallest number from the prrt region.
 Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
 	\centering
-	\includegraphics[width=\linewidth]{Figure4}
-	\label{fig4}
+	\includegraphics[width=6.5in]{FigureS2.emf}
 \end{figure}
 
-\begin{figure}[p!]
+\begin{figure}[H]
 	\caption{Clusters (vertical colored bars) inferred from the phylogenies of Sanger consensus sequences differ across genomic regions.
-The largest number of clusters was inferred from the int region, and the smallest number from the env region.
+The largest number of clusters was inferred from the env region, and the smallest number from the prrt region.
 Profile sampling provides a bootstrapped measure of cluster support (annotation to bars).}
 	\centering
-	\includegraphics[width=\linewidth]{Figure5}
-	\label{fig5}
-\end{figure}
-
-\newpage
-
-\begin{figure}[p!]
-	\caption{Summary of clusters identified by Sanger versus NGS consensus sequences across genomic regions.
-Numeric values indicate bootstrapped cluster support from the profile sampling method.
-A blank cell indicates that the cluster was not detected in that consensus sequence and genomic region.}
-	\centering
-	\includegraphics[width=3.25in]{Figure6}
-	\label{fig6}
+	\includegraphics[width=6.5in]{FigureS3.emf}
 \end{figure}
